---
title: ITS2 sequence analysis
author: "MP Martynek"
date: '2025'
output:
  html_document:
    toc: yes
    toc_depth: 6
    toc_float: yes
  pdf_document:
    keep_tex: yes
editor_options: 
  chunk_output_type: console
---

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

ITS2 strain variation data analysis based on scripts from Ariana Huffmyer, University of Washington and Shayle Matsuda, Shedd Aquarium.   

```{r}
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("readxl")) install.packages("readxl")
if (!require("phyloseq")) BiocManager::install("phyloseq")
if (!require("janitor")) install.packages("janitor")
if (!require("writexl")) install.packages("writexl")
#if (!require("microViz")) BiocManager::install("microViz")

library("readxl")
library("phyloseq")
library("janitor")
library("writexl")
library("ggplot2")
library('cowplot')
library("janitor")
library("car")
library("phyloseq")
library("stringr")
library("ape")
library("DESeq2")
#library("microViz")
library("vegan")
library("ComplexHeatmap")
library("tidyverse")
library("dplyr")
library(emmeans)
library(readxl)
```

# **Prepare data**  

Load metadata.  
```{r}
#create metadata and load
metadata <- read.csv("ITS2/SymPortal_datasheet_Martynek.csv", header = TRUE, sep = ",")%>%dplyr::select(sample_name, collection_depth, origin, timepoint)

remove <- c("26", "27", "43", "44","233", "234", "235", "236", "237", "238", "239", "240", "243", "244", "247", "248" )

metadata <- metadata %>%
  filter(!sample_name %in% remove)

rownames(metadata) <- metadata$sample_name

metadata <- as.matrix(metadata)
metadata <- sample_data(data.frame(metadata))
```

Load the data for strain level information.  

```{r}
#gather the taxonomy names  
taxnames <- read_tsv(
  file  = "ITS2/638_20251021T051401_DBV_20251021T231108.seqs.absolute.abund_only.txt",
  n_max = 0) %>%
  select(-1) %>%
  names(.)

#extract clade letter  
tax0 <- data_frame(
  DIV = taxnames,
  clade = str_extract(DIV, "[A-Z]")
)

tax1 <- as.matrix(tax0)
rownames(tax1) <- tax0$DIV
tax <- tax_table(tax1)

otu0 <- read_tsv(
  file  = "ITS2/638_20251021T051401_DBV_20251021T231108.seqs.absolute.abund_and_meta.txt") %>%
  select(-1, )

otu1 <- as.matrix(otu0[, 39:116])
rownames(otu1) <- otu0$sample_name
otu <- otu_table(otu1, taxa_are_rows = FALSE)

coralDIV <- phyloseq(otu, tax, metadata)
coralDIV
```

# **Build DIV Data**  

View the number of sequence counts in our dataset.   
```{r}
samps<-read_tsv(file  = "ITS2/638_20251021T051401_DBV_20251021T231108.seqs.absolute.abund_and_meta.txt")%>%select(post_taxa_id_absolute_symbiodiniaceae_seqs, sample_name)

hist(samps$post_taxa_id_absolute_symbiodiniaceae_seqs, breaks=20)

samps
```
We have >85K for most samples. DL03, SL10, FP04, FP09 which have about 60K. Note that these samples also had distinct type profiles from all other samples too 

Plot tree.  
```{r}
random_tree = rtree(ntaxa(coralDIV), rooted=TRUE, tip.label=taxa_names(coralDIV))
plot(random_tree)
```

Merge phyloseq data. 
```{r}
phylo_coral_DIV = merge_phyloseq(coralDIV, metadata, random_tree)
phylo_coral_DIV
```

Filter out samples with low numbers in groups (n=1 per group) from phyloseq object.  
```{r}
phylo_coral_DIV <- subset_samples(
  phylo_coral_DIV,
  !(sample_name %in% c("DL03", "SL10", "FP04", "FP09"))
)

sample_data(phylo_coral_DIV)$origin <- factor(
  sample_data(phylo_coral_DIV)$origin,
  levels = c("HB", "FP", "DL", "SL", "RC", "RF")  
)
```

We now have 173 samples. 

Now build a tree with the combined data.  
```{r}
plot_tree(phylo_coral_DIV, color="origin", label.tips="taxa_names", ladderize="left", plot.margin=0.3)
```

Plot a heatmap.  

```{r}
plot_heatmap(phylo_coral_DIV, "NMDS", "bray", "origin", low="#66CCFF", high="#000033", na.value="white")
```

Assign object to a shorter name. 

```{r}
GP1 = phylo_coral_DIV
GP1
```

Display smaple sequencing depth. 
```{r}
min(sample_sums(GP1))
```

Transform to relative abundance and subsample to even sampling depth to minimum count in dataset \
```{r}
GP1_rare<-rarefy_even_depth(GP1, sample.size = min(sample_sums(GP1)),
  rngseed = TRUE, replace = TRUE, trimOTUs = TRUE, verbose = TRUE)
```

View sample sequencing depth again after subsampling. 
```{r}
sample_sums(GP1_rare)
```

All samples are now at 58,487 sequences. 17 OTU's removed. 

Now transform counts to relative abundance expressed as a proportion.    

```{r}
#GP1_rare = transform_sample_counts(GP1_rare, function(x) 100 * x/sum(x))
GP1_rare = transform_sample_counts(GP1_rare, function(x) x/sum(x))
```

View tree and heatmap after subsampling.   
```{r}
plot_tree(GP1_rare, color="origin", label.tips="taxa_names", ladderize="left", plot.margin=0.3)
plot_heatmap(GP1_rare, "NMDS", "bray", "origin", low="#66CCFF", high="#000033", na.value="white")
```

# **DIV Ordinations: NMDS** 

View NMDS ordination by OTU.    
```{r}
GP.ord <- ordinate(GP1_rare, "NMDS", "bray")
ordinate(GP1_rare, "NMDS", "bray")
p1 = plot_ordination(GP1_rare, GP.ord, type="taxa", title="taxa")
p1
```

Plot with samples now.  
```{r}
p2 = plot_ordination(GP1_rare, GP.ord, type="samples", color="origin") 
p2_geom<- p2 + geom_polygon(aes(fill=origin)) + geom_point(size=4) + ggtitle("samples")
p2_geom
```

View PCoA by OTU.  
```{r}
GP.ordb.pcoa <- ordinate(GP1_rare, "PCoA", "bray")
p1b = plot_ordination(GP1_rare, GP.ordb.pcoa, type="taxa", title="taxa")
p1b
```

Plot with samples now.  
```{r}
p2b = plot_ordination(GP1_rare, GP.ord, type="samples", color="origin"); p2b 
```

Output NMDS plot for visualization.  
```{r}
Origin_names<-c("Harry's Bommie", "Fourth Point", "Deep Lagoon", "Shallow Lagoon", "Reef Crest", "Reef Flat")

p2b_viz<-p2  + 
  geom_point(aes(colour=origin), size=4) + 
  ggtitle("NMDS")+
  scale_colour_manual(name="Origin", values=c("#131142FF","dodgerblue2", "steelblue1", "brown1", "#C1000e", "violetred" ), labels=Origin_names) + ggtitle("A")+
  theme_classic() + 
  theme(
    panel.background=element_blank()
  ); p2b_viz

#ggsave(plot=p2b_viz, "figures/its2/nmds_its2_div.jpeg", h=5, w=6)
```

Conduct a permanova in the vegan package.  

```{r}
metadata <- as(sample_data(GP1_rare), "data.frame")

perm1<-adonis2(phyloseq::distance(GP1_rare, method="bray") ~ origin,
       data = metadata)

perm1
```

Permutation test for adonis under reduced model
Permutation: free
Number of permutations: 999

adonis2(formula = phyloseq::distance(GP1_rare, method = "bray") ~ origin, data = metadata)
          Df SumOfSqs      R2      F Pr(>F)    
Model      5  0.40171 0.17875 7.2699  0.001 ***
Residual 167  1.84555 0.82125                  
Total    172  2.24726 1.00000                  
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1   

# **Plot heat map of ITS2 DIVs**  

Draw with Phyloseq.  

Filter out taxa that comprise <1% mean relative abundance across samples. 

There were 61 OTU's before filtering and 16 after filtering. This indicates there were many rare taxa.   
```{r}
GP1_rare

GP1_prune = filter_taxa(GP1_rare, function(x) mean(x) > 0.01, TRUE)
GP1_prune 

#sample_order<-c("R55", "R56", "R57", "R58", "R59", "R60", "R61", "R62", "R63", "R64", "R65", "R66", "R67", "R68", "R69", "R71", "R72")

p<-plot_heatmap(GP1_prune, sample.label="origin", low="#66CCFF", high="#000033", na.value="white", taxa.label="DIV")
p$scales$scales[[1]]$name <- "Origin"

print(p)

#ggplot2::ggsave("Mcap2020/Figures/ITS2/div_heatmap.jpg", p, width = 6, height = 6)
```

Run the NMDS visualization again with the dataset with rare samples removed. Run a PERMANOVA.  

```{r}
GP.ordb.filt <- ordinate(GP1_prune, "NMDS", "bray")

p3b = plot_ordination(GP1_prune, GP.ordb.filt, type="samples", color="origin"); p3b 

p3b_viz_filt<-p3b  + 
  geom_point(aes(colour=origin), size=4) + 
  ggtitle("NMDS")+
  scale_colour_manual(name="Origin", values=c("#131142FF","dodgerblue2", "steelblue1", "brown1", "#C1000e", "violetred" ), labels=Origin_names)+
  theme_classic() + 
  ggtitle("A")+
  theme(
    panel.background=element_blank()
  ); p3b_viz_filt

#ggsave(plot=p3b_viz_filt, "figures/its2/nmds_its2_div_filtered.jpeg", h=5, w=6)
```

Run PERMANOVA on filtered data set. 
```{r}
metadata_filt <- as(sample_data(GP1_prune), "data.frame")

perm2_filt<-adonis2(phyloseq::distance(GP1_prune, method="bray") ~ origin,
       data = metadata_filt)

perm2_filt
```

Permutation test for adonis under reduced model
Permutation: free
Number of permutations: 999

adonis2(formula = phyloseq::distance(GP1_prune, method = "bray") ~ origin, data = metadata_filt)
          Df SumOfSqs      R2      F Pr(>F)    
Model      5  0.14247 0.12681 4.8506  0.001 ***
Residual 167  0.98102 0.87319                  
Total    172  1.12349 1.00000                  
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Draw by origin ih Complex Heatmap.  
```{r}
# library(dendsort)
# library(WGCNA)
# library(circlize)
# 
# table<-as.matrix(t(otu_table(GP1_prune)))
#                  
# row_dend = dendsort(hclust(dist(table)))
# col_dend = dendsort(hclust(dist(t(table))))
# 
# #col_fun = colorRamp2(c(0, 50, 100), c("white", "darkgray", "black"))
# col_fun = colorRamp2(c(0, .1, .25, .5), hcl_palette = "Blue-Red 3")
# 
# #pdf(file = "figures/its2/DIV-ComplexHeatmap.pdf", height = 7, width = 7)
# ComplexHeatmap::Heatmap(table, name = "Relative \nAbundance", row_title = "DIV", column_title = "ITS2 DIV Relative Abundance (>1%)", 
#         col = col_fun, 
#         row_names_side = "left", 
#         row_dend_side = "left",
#         width = unit(4, "in"), 
#         height = unit(4, "in"), 
#         column_dend_height = unit(.5, "in"),
#         row_dend_width = unit(.5, "in"),
#         column_dend_reorder = FALSE,
#         row_dend_reorder = TRUE,
#         row_gap = unit(2.5, "mm"),
#         clustering_distance_rows = "euclidean",
#         clustering_method_rows = "complete",
#         border = TRUE, 
#         column_names_gp =  gpar(fontsize = 12, border=FALSE),
#         column_names_rot = 35,
#         cluster_rows = row_dend, 
#         cluster_columns = col_dend,
#         column_order = sample_order, 
#         row_names_gp = gpar(fontsize = 12, alpha = 0.75, border = FALSE))
# dev.off()

```

# **Alpha Diversity**  

Calculate diversity of those that have more than 1 sequence (no singletons).
```{r}
GP1_rare_untransformed<-rarefy_even_depth(GP1, sample.size = min(sample_sums(GP1)),
  rngseed = TRUE, replace = TRUE, trimOTUs = TRUE, verbose = TRUE)

GP2 <- prune_taxa(taxa_sums(GP1_rare_untransformed) > 1, GP1_rare_untransformed)
```

Plot all richness metrics.  
```{r}
plot_richness(GP2)
```

Plot the two most commonly used - Shannon and Inverse Simpson.  
```{r}
alpha_df <- estimate_richness(GP2, measures = c("InvSimpson", "Shannon"))

# 2. Add origin info
alpha_df$origin <- sample_data(GP2)$origin
alpha_df$timepoint <- factor(sample_data(GP2)$timepoint,
                             levels = c("Sept_2022", "May_2024", "Sept_2024"))

# 3. Pivot longer so we have a column "variable" and "value"
alpha_long <- alpha_df %>%
  pivot_longer(cols = c("InvSimpson", "Shannon"),
               names_to = "variable",
               values_to = "value")

# 4. Plot
alpha1_plot <- ggplot(alpha_long, aes(x = origin, y = value, color = origin, shape = timepoint)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.6), size = 3, alpha = 0.8) +
  facet_wrap(~variable, scales = "free_y") +
  theme_classic() +
  scale_color_manual(values=c("#131142FF","dodgerblue2","steelblue1",
                              "brown1","#C1000e","violetred")) +
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"),
        axis.title.x = element_blank())

alpha1_plot

#ggsave(plot=alpha1_plot, file="figures/its2/alpha_rare.png", width=7, height=5)
```

Test for differences in alpha diversity.  
```{r}
metadata <- as(sample_data(GP2), "data.frame")
metadata$sample<-rownames(metadata)

alpha_rare<-estimate_richness(GP2, measures=c("InvSimpson", "Shannon", "Chao1"))
alpha_rare$sample<-rownames(alpha_rare)

alpha_rare<-full_join(metadata, alpha_rare, by="sample")

summary(aov(Shannon~origin, data=alpha_rare))
summary(aov(InvSimpson~origin, data=alpha_rare))
summary(aov(Chao1~origin, data=alpha_rare))
```


# **Plot profile level information** 

Load profile level data.  
```{r}
#create metadata and load
metadata <- read.csv("ITS2/SymPortal_datasheet_Martynek.csv", header = TRUE, sep = ",")%>%dplyr::select(sample_name, collection_depth, origin, timepoint)

remove <- c("26", "27", "43", "44","233", "234", "235", "236", "237", "238", "239", "240", "243", "244", "247", "248" )

metadata <- metadata %>%
  filter(!sample_name %in% remove)

rownames(metadata) <- metadata$sample_name

metadata <- as.matrix(metadata)
metadata <- sample_data(data.frame(metadata))

#load data for taxonomy
tax0 <- read_tsv(
  file  = "ITS2/638_20251021T051401_DBV_20251021T231108.profiles.absolute.abund_and_meta.txt", n_max = 6) %>%
  dplyr::select(-2) %>% 
  gather(uid, value, -1) %>% 
  spread(1, value) %>%
  clean_names()

tax1 <- as.matrix(tax0[, -1], dimnames = list(tax0$uid, colnames(tax0[-1])))
rownames(tax1) <- tax0$uid
tax <- tax_table(tax1)

#load data for count matrix 
otu0 <- read_tsv(
  file  = "ITS2/638_20251021T051401_DBV_20251021T231108.profiles.absolute.abund_and_meta.txt", col_names = TRUE) 

list<-c("UID", "sample_name", "1585934", "1586098", "1587534", "1586603", "1585960", "1585391", "1586361", "1587514", "1586753", "1588633", "1588394", "1585393", "1585933", "1587213", "1586587", "1589029", "1586365", "1587322", "1591001", "1590984")
colnames(otu0)<-list

otu0<-otu0%>%
  select(-1) %>%
  #dplyr::slice(7:24)
  mutate_at(2:ncol(.), as.numeric)

otu1 <- as.matrix(otu0[, -1])
rownames(otu1) <- otu0$sample_name
otu <- otu_table(otu1, taxa_are_rows = FALSE)

taxa_names(otu)
taxa_names(tax)

#combine to phyloseq object
coral <- phyloseq(otu, tax, metadata)
coral
```

Create tree.  
```{r}
random_tree_profile = rtree(ntaxa(coral), rooted=TRUE, tip.label=taxa_names(coral))
```

Merge phyloseq data. 
```{r}
phylo_coral = merge_phyloseq(coral, metadata, random_tree_profile)
phylo_coral
```

Filter out samples with low numbers in groups from phyloseq object.  
```{r}
phylo_coral_DIV <- subset_samples(
  phylo_coral_DIV,
  !(sample_name %in% c("DL03", "SL10", "FP04", "FP09"))
)

sample_data(phylo_coral_DIV)$origin <- factor(
  sample_data(phylo_coral_DIV)$origin,
  levels = c("HB", "FP", "DL", "SL", "RC", "RF")  
)
```

Now build a tree with the combined data.  
```{r}
plot_tree(phylo_coral, color="origin", label.tips="taxa_names", ladderize="left", plot.margin=0.3)
```

Plot a heatmap.  

```{r}
plot_heatmap(phylo_coral, "NMDS", "bray", "origin", low="#66CCFF", high="#000033", na.value="white")
```

Plot ordinations.  

```{r}
GP1_profile = phylo_coral
GP1_profile
```


```{r}
min(sample_sums(GP1_profile))

GP1_rare_profile<-rarefy_even_depth(GP1_profile, sample.size = min(sample_sums(GP1_profile)),
  rngseed = TRUE, replace = TRUE, trimOTUs = TRUE, verbose = TRUE)

sample_sums(GP1_rare_profile)
```
All samples now have 38483 sequences. 

Transform to relative abundance as a proportion. 
```{r}
GP1_rare_profile = transform_sample_counts(GP1_rare_profile, function(x) x/sum(x))

GP1_profile = transform_sample_counts(GP1_profile, function(x) x/sum(x))
```

View tree and heatmap after subsampling  
```{r}
plot_tree(GP1_rare_profile, color="origin", label.tips="taxa_names", ladderize="left", plot.margin=0.3)
plot_heatmap(GP1_rare_profile, "NMDS", "bray", "origin", low="#66CCFF", high="#000033", na.value="white")
```

Plot abundance of each profile across origins
#Fig 4
```{r}

library(wesanderson)

GP1_S22 <- subset_samples(GP1_profile, timepoint == "Sept_2022")

samples_to_remove <- c("DL06", "FP04")

GP1_S22 <- subset_samples(GP1_S22, !(sample_names(GP1_S22) %in% samples_to_remove))

GP1_S22 <- prune_taxa(taxa_sums(GP1_S22) > 0, GP1_S22)


sample_data(GP1_S22)$origin <- factor(
  sample_data(GP1_S22)$origin,
  levels = c("HB", "FP", "DL", "SL", "RC", "RF")
)

Origin_names <- c(HB = "Harry’s Bommie",
                  FP = "Fourth Point",
                  DL = "Deep Lagoon",
                  SL = "Shallow Lagoon",
                  RC = "Reef Crest",
                  RF = "Reef Flat")

#its2_levels <- sort(unique(as.character(t$data$its2_type_profile)))

#its2_levels

my_its2_cols <- c(
  "C42a/C1/C42.2/C1b/C3-C1j-C1au-C115k" = "firebrick2",
  "C42a/C42aw/C1-C42.2-C1b-C1j-C1au-C3" = "steelblue",
  "C1/C42a/C42.2/C1b/C1j-C1au-C3-C115l" = "blue",
  "C42a/C1/C42.2/C1b/C1j-C1ci-C1au-C3"  = "turquoise1",
  "C42a/C42.2/C1-C1au-C1b-C3-C115k"     = "royalblue3",
  "C42dh/C1-C42.2-C1b-C42dz"            = "darkslategray4",
  "C15/C93a"                            = "deepskyblue2"
)

t<-plot_bar(GP1_S22, fill="its2_type_profile")+ geom_bar(stat="identity", width = 0.9)

t_plot<-t  + facet_grid(~origin, scales = "free", space = "free", labeller = labeller(origin=Origin_names)) + ylab("Relative Abundance (%)") + theme(strip.text.x = element_text(size = 5, face="bold"))+
  theme_classic() +
  scale_fill_manual(values = my_its2_cols,breaks = names(my_its2_cols), name = "ITS2 Type Profile", drop = TRUE)+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size = 16), axis.title.y = element_text(size = 16), axis.title.x = element_text(size = 16, face = "bold"), axis.text.y = element_text(size=16),legend.title = element_text(size = 16, face = "bold"), strip.text = element_text(size = 16, face = "bold"),  legend.text = element_text(size = 12)); t_plot

rel_abund_data<-t$data


```

```{r}

library(tidyverse)
library(scales)

# read files
abund <- read.delim("ITS2/20251021T032609_martyma1/post_med_seqs/638_20251021T051401_DBV_20251021T231108.seqs.relative.abund_only.txt",
                    check.names = FALSE)
meta  <- read.delim("ITS2/20251021T032609_martyma1/post_med_seqs/638_20251021T051401_DBV_20251021T231108.seqs.relative.meta_only.txt",
                    check.names = FALSE)

# join metadata to abundance table
its2 <- left_join(meta, abund, by = "sample_uid")

# keep Pocillopora damicornis Sept 2022 samples
# here Sept 2022 is represented by collection_date == 20220901 in the metadata
its2_s22 <- its2 %>%
  filter(host_genus == "pocillopora",
         host_species == "damicornis",
         collection_date == 20220901)

# remove samples you said to exclude
its2_s22 <- its2_s22 %>%
  filter(!sample_name %in% c("DL06", "FP04"))

# identify the sequence columns
seq_cols <- setdiff(colnames(abund), "sample_uid")

# pivot longer
its2_long <- its2_s22 %>%
  select(sample_uid, sample_name, collection_date, collection_depth, all_of(seq_cols)) %>%
  pivot_longer(cols = all_of(seq_cols),
               names_to = "ITS2_type",
               values_to = "RelAbund")

# make site column from sample name prefix
its2_long <- its2_long %>%
  mutate(Origin = str_extract(sample_name, "^[A-Z]+"))

# collapse anything <5% within each sample into Other
its2_plot <- its2_long %>%
  mutate(ITS2_plot = if_else(RelAbund < 0.05, "Other", ITS2_type)) %>%
  group_by(sample_name, Origin, ITS2_plot) %>%
  dplyr::summarise(RelAbund = sum(RelAbund), .groups = "drop")

tax_order <- its2_plot %>%
  group_by(ITS2_plot) %>%
  dplyr::summarise(total = sum(RelAbund), .groups = "drop") %>%
  dplyr::arrange(desc(total)) %>%
  pull(ITS2_plot)

tax_order <- c(setdiff(tax_order, "Other"), "Other")

its2_plot <- its2_plot %>%
  mutate(
    Origin = factor(Origin, levels = c("HB", "FP", "DL", "SL", "RC", "RF")),
    ITS2_plot = factor(ITS2_plot, levels = tax_order)
  )

# palette with Other as gray
taxa_levels <- levels(its2_plot$ITS2_plot)
blue_taxa <- setdiff(taxa_levels, "Other")
blue_cols <- colorRampPalette(c("#c6dbef", "#6baed6", "#2171b5", "#08306b"))(length(blue_taxa))
pal <- setNames(c(blue_cols, "gray70"), c(blue_taxa, "Other"))
pal <- pal[taxa_levels]

# override specific sequence colors
pal[grepl("C3", names(pal))]  <- "deeppink"
pal[grepl("C15", names(pal))] <- "darkorange"
pal[grepl("C42aw", names(pal))] <- "springgreen4"
pal[grepl("C115k", names(pal))] <- "yellow2"
pal[grepl("C42dh", names(pal))] <- "springgreen4"
pal[grepl("C1hr", names(pal))] <- "magenta4"
pal[grepl("19017_C", names(pal))] <- "brown"
pal[grepl("52544_C", names(pal))] <- "darkolivegreen"
pal[grepl("C1fi", names(pal))] <- "sienna3"


# plot
Fig4 <- ggplot(its2_plot, aes(x = sample_name, y = RelAbund, fill = ITS2_plot)) +
  geom_col(width = 0.85) +
  facet_grid(~Origin, scales = "free_x", space = "free_x") +
   scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = c(0, 0)
  )  +
  scale_fill_manual(values = pal, name = "ITS2 sequence") +
  labs(x = expression(bold("Sample")),
     y = expression(bold("Relative abundance")))+
  theme_classic() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 16), axis.title.y = element_text(size = 16), axis.title.x = element_text(size = 16), axis.text.y = element_text(size=16),legend.title = element_text(size = 16, face = "bold"), strip.text = element_text(size = 16, face = "bold"), legend.text = element_text(size = 12))

Fig4



```


```{r}
library(patchwork)

Fig4_combined <- (
  (Fig4 + 
     labs(x = NULL) + 
     theme(
       axis.text.x = element_blank(),
       axis.ticks.x = element_blank()
     )) /
  (t_plot + 
     labs(y = NULL) + 
     theme(
       axis.title.y = element_blank(),
       strip.text = element_blank()
     )) +
  plot_layout(heights = c(5, 1.2), guides = "collect")
) +
  plot_annotation(
    tag_levels = "A",
    theme = theme(
      plot.tag = element_text(size = 20, face = "bold"),
      plot.tag.position = c(0, 1)
    )
  ) &
  theme(legend.position = "right")

Fig4_combined <- Fig4_combined & theme(
  plot.tag = element_text(size = 18, face = "bold")
)

Fig4_combined
```

```{r}

ggsave("FIG4_V2_combined.pdf",
       plot = Fig4_combined,
       width = 16,       # try 10–14 inches depending on how many columns
       height = 8,       # adjust height accordingly
       dpi = 600,        # high resolution for publication
       units = "in")     # use inches for precise control



```



```{r}

its2_sept22 <- subset_samples(GP1_rare_profile, timepoint == "Sept_2022")


ord_sept22 <- ordinate(GP1_S22, method = "NMDS", distance = "bray")

p_nmds <- plot_ordination(its2_sept22, ord_sept22, color = "origin") +
  geom_point(size = 3, alpha = 0.8) +
  theme_classic() +
  ggtitle("NMDS - Sept 2022")

p_nmds

```

PERMANOVA
```{r}
library(pairwiseAdonis)


metadata_s22 <- as(sample_data(GP1_S22), "data.frame")

perm2<-adonis2(phyloseq::distance(GP1_S22, method="bray") ~ origin,
       data = metadata_s22)

perm2

dist_sept22 <- phyloseq::distance(GP1_S22, method = "bray")

# Now run pairwise
pairwise_results <- pairwise.adonis(
  dist_sept22,
  factors = metadata_s22$origin,
  p.adjust.m = "bonferroni"
)

pairwise_results


```


Calculate the average percent abundance for each taxa in each origin  
```{r}
means <- rel_abund_data %>%
  group_by(its2_type_profile, origin) %>%
  # Add abundances within each profile and origin
  summarize_at("Abundance", mean)%>%
  spread(its2_type_profile, Abundance);means
```


```{r}
library(broom)

aov_results <- rel_abund_data %>%
  group_by(its2_type_profile) %>%
  do(tidy(aov(Abundance ~ origin, data = .)))

aov_results
```


