---
title: "Reef-scale environmental mosaics shape coral physiological and microbial signatures
  of resilience"
author: "Marcelina Martynek"
date: "2026-05-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***Initial sequence processing in DADA2 was performed on a larger dataset that included the September 2022 samples analyzed in this manuscript together with additional samples not considered further here. Downstream analyses for this study were then conducted using only the September 2022 samples.***


#working nonsense
```{r}
library(dada2); packageVersion("dada2") # 1.16 or later
library(ggplot2)
library(ShortRead)
library(Biostrings)
library(stringr)
library(readr)
library(dplyr)
library(phyloseq)
library(decontam)
library(microbiome)
library(ggpubr)
library(multcompView)
library(DECIPHER)
library(phangorn)
library(ape)
library(vegan)
library(Rmisc)
library(tidyverse)
library(RColorBrewer)
library(patchwork)
library(pairwiseAdonis)
#library(MicEco)
library(DESeq2)
library(RColorBrewer)
library(Hmisc)

```

```{r}
ps <- readRDS("ps_tree.rds")

# Optional sanity check — will throw error if tree is missing
phy_tree(ps)

DNA_concentration<-read.csv("Barott Coral Extractions - Marcelina - Sheet1.csv", strip.white=T)
DNA_concentration$Origin<-as.factor(DNA_concentration$Origin)
DNA_concentration$SampleID<-as.factor(DNA_concentration$SampleID)

path <- "C:/Users/marce/Desktop/Heron/CHOP_Heron_Jan2025/demultiplexed_fastq"
list.files(path, all.files = TRUE)

```

#Pre-sequencing: DNA concentration visualization
```{r}
DNA_concentration$Origin<- factor(DNA_concentration$Origin, levels =c("RF","RC","SL","DL", "FP", "HB", "Control")) 

DNA_concentration$study_group <- factor(DNA_concentration$study_group, 
                                         levels = c("Sept_2022", "May_2024", "Sept_2024"))
concentration_plot <- ggplot(DNA_concentration, aes(x = study_group, y = DNA_concentration_ng_uL)) + 
  geom_point(aes(color = Origin), position = position_jitter(width = 0.05), alpha = 0.8) +  
  theme_classic() + 
  theme(axis.text.x = element_text(vjust = 0.5, size = 12, angle = 45), # Angled x-axis labels
        axis.text.y = element_text(vjust = 0.5, size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        legend.text = element_text(vjust = 0.5, size = 12),
        panel.background = element_rect(fill = NA, color = 'black'),
        strip.text = element_text(size = 12, face = "italic")) +
  facet_wrap(~ Origin, scales = "free_y", ncol = 7) + coord_cartesian(ylim = c(0, 30))

concentration_plot

```

#Fasta-file processing 
```{r}

head(list.files(path))

# Read in the forward and reverse fastq file names
fnFs <- list.files(path, pattern="_R1.fastq.gz", full.names=TRUE)
fnRs <- list.files(path, pattern="_R2.fastq.gz", full.names=TRUE)
head(fnFs)


samples = str_extract(basename(fnFs), "^[^_]+")


names(fnFs) <- samples
names(fnRs) <- samples
```
```{r}

fwd_primer <- "GTGCCAGCAGCCGCGGTAA"  #Wildcards: Y (4th position) = C or T ; M (9th position) = A or C
rev_primer <- "GGACTACACGGGTWTCTAAT" #WCs: N (8th position) = G or A or T or C ; V (9th position) = A or C or G
fwd_primer_rev <- as.character(reverseComplement(DNAStringSet(fwd_primer))) #can't get the reverse to work 
rev_primer_rev <- as.character(reverseComplement(DNAStringSet(rev_primer)))

# This function counts number of reads in which the primer is found
count_primers <- function(primer, filename) {
  num_hits <- vcountPattern(primer, sread(readFastq(filename)), fixed = FALSE)
  return(sum(num_hits > 0))
}

count_primers(fwd_primer, fnFs[[1]])
count_primers(rev_primer, fnRs[[1]])

#Sanity check complete! no primers whatsoever 

```
```{r}
plotQualityProfile(fnFs[c(2:10)])
plotQualityProfile(fnRs[c(2:10)])
# Good quality/bad quality? What are the read lengths?
# How might these quality profiles inform our choice of truncation lengths?
```
```{r}
# Define the paths to filtered files we are going to create
filtFs <- file.path(path, "filtered", basename(fnFs)) 
filtRs <- file.path(path, "filtered", basename(fnRs))
# The filtered files will be in the `filtered/` sub-directory within `path`

```

#Filter and Trim 
```{r}

# Perform filtering and trimming (filterandTrim doesn't work for multithreading with Windows)
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, maxEE=5, 
                     trimLeft=c(10,0), truncLen=c(240,200), truncQ=4) # REPLACE XXX/YYY with valid parameter choices
                     # REPLACE XXX/YYY with valid parameter choices
out
```
```{r}

filtFs<- filtFs[file.exists(filtFs)] #filter out S9 because that sample failed
filtRs <- filtRs[file.exists(filtRs)]

# Learn the error model from the filtered data.
errF <- learnErrors(filtFs, multi=TRUE) # `multi=TRUE` activates multi-threading to reduce computation time
errR <- learnErrors(filtRs, multi=TRUE)

# Visualize the error model. Points are observations, black line is fitted error model.`
plotErrors(errF)
plotErrors(errR)
# Do the fitted error models look reasonable?

# Run the DADA2 method using the fitted error model.
ddF <- dada(filtFs, errF, pool="pseudo", multi=TRUE)
ddR <- dada(filtRs, errR, pool="pseudo", multi=TRUE)
# What pooling option makes sense? FALSE (default), "pseudo", or TRUE? See ?dada for more info
# For more pseudo-pooling detail: https://benjjneb.github.io/dada2/pseudo.html#pseudo-pooling
# Challenge: Try different pooling options and compare them.

# Merge the denoised forward and reverse reads together.
mm <- mergePairs(ddF, filtFs, ddR, filtRs, verbose=TRUE)

# Construct a sequence table: rows are samples, columns are ASVs, values are abundances.
# ASVs (columns) are ordered by their summed abundance across all samples.
sta <- makeSequenceTable(mm)
dim(sta)


# Remove chimeric ASVs and construct a new chimera-free sequence table.
st <- removeBimeraDenovo(sta, multi=TRUE, verbose=TRUE)
sum(st)/sum(sta)
# Were most reads retained during chimera removal? How about ASVs? 
# How does the fraction of chimeric ASVs compare to the fraction of chimeric reads?

##inspect the numbers of reads


# In this case, most reads should make it through the entire pipeline!
# Most importantly, a large majority (>80% of reads) should merge successfully,
#  and almost all (>95%)  reads should pass chimera filtering.
# IF THAT ISN'T THE CASE, you have a problem, and need to revisit your truncation lengths
#   (merging problem) or primer removal (trimLeft, chimera problem).

# Code derived from the dada2 tutorial: https://benjjneb.github.io/dada2/tutorial.html
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(ddF, getN), sapply(ddR, getN), sapply(mm, getN), rowSums(st))
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- basename(fnFs)
head(track)  #reads look good!


```

#Assign taxonomy 
```{r}
# The dada2 taxonomic reference page https://benjjneb.github.io/dada2/training.html has links to a 
#   number of reference databases formatted to work with `assignTaxonomy`.I recommend Silva for 16S.
#   However, here we will use RDP in this lab for file size reasons.
#download.file("https://zenodo.org/record/4310151/files/rdp_train_set_18.fa.gz?download=1",
#              "rdp_train_set_18.fa.gz"


# Assign taxonomy down to the genus level to these 16S ASVs.
tax <- assignTaxonomy(st, "C:/Users/marce/Desktop/Heron/CHOP_Files/silva_nr99_v138.1_wSpecies_train_set.fa.gz", multi=TRUE)
unname(head(tax)) # unname strips the long ASV rownames to make the printed output more readable
# Are reasonable taxonomic assignments being made to the abundant taxa? YES

addSpecies <- addSpecies(tax, "C:/Users/marce/Desktop/Heron/CHOP_Files/silva_species_assignment_v138.1.fa")

write.table(addSpecies, "C:/Users/marce/Desktop/Heron/CHOP_Heron_Jan2025/16S_taxonomy.csv", row.names = TRUE, sep ="\t", quote =FALSE)

```

#Let's check on the Mock community
```{r}
# unqs.mock <- st["mockdna1_R1.fastq.gz",]
# unqs.mock <- sort(unqs.mock[unqs.mock>0], decreasing=TRUE) # Drop ASVs absent in the Mock
# cat("DADA2 inferred", length(unqs.mock), "sample sequences present in the Mock community.\n")
# 
# mock.ref <- getSequences(file.path(path, "mockdna1.fasta"))
# match.ref <- sum(sapply(names(unqs.mock), function(x) any(grepl(x, mock.ref))))
# cat("Of those,", sum(match.ref), "were exact matches to the expected reference sequences.\n")

```

#Constuct Phylogentic Tree 
```{r}
sequences<-getSequences(st) #from DADA2
names(sequences)<-sequences
```
```{r}
# Align sequences
alignment <- AlignSeqs(DNAStringSet(sequences), anchor = NA)

alignment_matrix <- as.matrix(alignment)
phang.align <- phyDat(alignment_matrix, type = "DNA")

# Build distance matrix and tree
dm <- dist.ml(phang.align)
treeNJ <- NJ(dm)

# Optional: optimize tree
fit <- pml(treeNJ, data = phang.align)

```
```{r}

fitGTR <- update(fit, k=4, inv=0.2) #calculate tree likelihood and optimize model
#fitGTR <- optim.pml(fitGTR, model="GTR", optInv=TRUE, optGamma=TRUE,
                    #rearrangement = "stochastic", control = pml.control(trace = 0))  #take WAY TOO LONG - aborted 

save(fitGTR, file="fitGTR.RData")

tree <- fitGTR$tree


ps@phy_tree<-fitGTR$tree
saveRDS(ps, "CHOP_Heron_Jan2025/ps_tree.rds")
ps <- readRDS("CHOP_Heron_Jan2025/ps_tree.rds")

ps


```

#Read in the metadata
```{r}
metadata <- read.csv("C:/Users/marce/Desktop/Heron/CHOP_Heron_Jan2025/CHOPMC430_Barrot_only_16S_mapping_file.tsv", 
                     header = TRUE,
                     check.names = FALSE)

# Save it as a clean CSV
write.csv(metadata, 
          "C:/Users/marce/Desktop/Heron/CHOP_Heron_Jan2025/metadata.csv", 
          row.names = FALSE)

#create new metadata file with relevant data 
Heron16S_full <- read.csv("C:/Users/marce/Desktop/Heron/CHOP_Heron_Jan2025/Heron16S_metadata_full.csv")

#make rownames 
Heron16S_full <- data.frame(Heron16S_full[,-1], row.names=Heron16S_full[,1])

#create phyloseq object 

#ps<- readRDS("CHOP_Heron_Jan2025/ps.rds")

ps <- phyloseq(sample_data(Heron16S_full),otu_table(st, taxa_are_rows=FALSE), tax_table(addSpecies), tree)
#ps <- phyloseq(sample_data(Heron16S_full),otu_table(st, taxa_are_rows=FALSE), tax_table(addSpecies))

ps

```
##root the tree 
```{r}

set.seed(711)
phy_tree(ps) <- root(phy_tree(ps), sample(taxa_names(ps), 1), resolve.root = TRUE)
is.rooted(phy_tree(ps))

```

##replace w/ ASVs 
```{r}

dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))

saveRDS(ps, "CHOP_Heron_Jan2025/ps_tree.rds")

ps
# ^^^ add ASVs: more convenient to use short names for our ASVs (e.g. ASV21) rather than 
#the full DNA sequence when working with some of the tables and visualizations from 
#phyloseq, but we want to keep the full DNA sequences for other purposes like merging 
#with other datasets or indexing into reference databases like the Earth Microbiome Project. 
#For that reason we’ll store the DNA sequences of our ASVs in the refseq slot of the phyloseq object, 
#and then rename our taxa to a short string. That way, the short new taxa names will appear in tables and plots, 
#and we can still recover the DNA sequences corresponding to each ASV as needed with refseq(ps). 
```


#Decontam 
```{r}
library(decontam); packageVersion("decontam")
download.file("https://raw.githubusercontent.com/mblstamps/stamps2024/main/dada2-and-decontam-labs/decontam_lab_bile.R", "decontam_lab_bile.R")

head(sample_data(ps))
sample_data(ps)$controls_or_naw <- ifelse(sample_data(ps)$Sample_Type == "Control", T, F) #create vector

df <- data.frame(sample_data(ps)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(ps)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize, color=Sample_Type)) + geom_point()

#sample_data(ps)$SampleID<-rownames( sample_data(ps))
sample_data(ps)$final_library_conc_ng_ul<- as.numeric(sample_data(ps)$final_library_conc_ng_ul)

```

##Combined method
```{r}

contamdf.freq <- isContaminant(ps, method="combined", conc="final_library_conc_ng_ul", neg="controls_or_naw", threshold=0.1)
head(contamdf.freq)
table(contamdf.freq$contaminant) #returned 174 TRUE contaminants
head(which(contamdf.freq$contaminant))  

#visualize between a "FALSE" ASV and a "TRUE" ASV 
plot_frequency(ps, taxa_names(ps)[c(1,23)], conc="final_library_conc_ng_ul") + 
  xlab("DNA Concentration (Final Library prep")

#let's look at a couple more classfied "TRUE"s
set.seed(100)
plot_frequency(ps, taxa_names(ps)[sample(which(contamdf.freq$contaminant),3)], conc="final_library_conc_ng_ul") +
  xlab("DNA Concentration")
```

```{r}

# Inspecting sequences at the extremes of the score distribution can help guide
#   choices about contaminant classification.
# Let's look more closely at the 10 highest scores (the 10 ASVs that look the most like non-contaminants according to decontam)
i.top10 <- order(contamdf.freq$p, decreasing=TRUE)[1:10]
contamdf.freq[i.top10,]
# Print out those ASV sequences in a fasta format

#dada2:::pfasta(sq[contamdf.freq$ASV[i.top10]], id=contamdf.freq$Phylum[i.top10])
# Use the NCBI web BLAST interface (https://blast.ncbi.nlm.nih.gov/Blast.cgi?PROGRAM=blastn&BLAST_SPEC=GeoBlast&PAGE_TYPE=BlastSearch) 
# to BLAST those sequences against Standard databases (nr etc.) -- nr/nt


# Now look at the the 10 lowest scores (the 10 ASVs that look the most like contaminants according to decontam)
i.bottom10 <- order(contamdf.freq$p, decreasing=FALSE)[1:10]
contamdf.freq[i.bottom10,] #9/10 classified as contaminates
#dada2:::pfasta(sq[contam$ASV[i.bottom10]], id=contam$Phylum[i.bottom10])

```
##Prune contaminants
```{r}
#get rid of contaminants!
ps.clean <- prune_taxa(!contamdf.freq$contaminant, ps)
ps.clean #okay so ps originally had 11953 taxa and cleaning the contaminants made it go down to 11779 taxa - so almost 200 taxa removed

saveRDS(ps.clean, "CHOP_Heron_Jan2025/ps.cleantree.rds")

#ps.clean<-readRDS('CHOP_Files/ps.cleantree.rds')


#plot # of reads by ASVs
ASV_sums<-as.data.frame(taxa_sums(ps))
plot_bar(ASV_sums) 

```

##Remove global singletons 
```{r}
#decontam takes care of a lot of contaminants first but just to ensure

sum(taxa_sums(ps.clean) == 1)
# just 1 

ps.clean <- prune_taxa(taxa_sums(ps.clean) > 1, ps.clean) #now 11768 taxa over 187 samples 


```

#Initial Exploration of data 
```{r}

qplot(log10(rowSums(otu_table(ps.clean))),binwidth=0.2) +
  xlab("Logged counts-per-sample") 

rel_abund <- t(apply(otu_table(ps.clean), 1, function(x) x / sum(x)))
qplot(rel_abund[, 12], geom = "histogram",binwidth=0.05) +
  xlab("Relative abundance")
```
```{r}
nsamples(ps.clean)
sample_names(ps.clean)
sample_variables(ps.clean)

# check out our otu count data 
sample_sums(ps.clean)
ntaxa(ps.clean)
head(taxa_sums(ps.clean), 20)

# check out our taxonomy table + filtering 
rank_names(ps.clean)
table(tax_table(ps.clean)[, "Class"], exclude = NULL)

```

#Filtering: take out chloroplast + mitochondria 
```{r}
ps.clean <- subset_taxa(ps.clean, Family != "Mitochondria") #ps.clean 
ps.clean <- subset_taxa(ps.clean, Order != "Chloroplast") 

#now 8839 taxa! 

saveRDS(ps.clean, 'CHOP_Heron_Jan2025/ps.cleantree.rds')

```

##Only P.dam 
```{r}
ps_pdam <- subset_samples(ps.clean, Type_ID %in% c("Pocillopora damicornis"))
ps_pdam

#ps_mock <-subset_samples(ps.clean, Sample_Type %in% c("Real", "MockDNA"))

otu <- as.data.frame(otu_table(ps_pdam))
tax <- as.data.frame(tax_table(ps_pdam))

#  transpose if samples are in columns instead of rows
if (taxa_are_rows(ps_pdam)) {
  otu <- t(otu_pdam)
}

# Create a vector of taxa names from the taxonomy table
# You can customize how the names are built — for example, using Genus and Species
tax_names <- apply(tax, 1, function(x) {
  paste(na.omit(x), collapse = "_")
})

# Replace ASV IDs with the constructed taxa names
colnames(otu) <- tax_names[colnames(otu)]

# Check the updated otu table
head(otu)


```

##filter samples
```{r}
sum(sample_sums(ps_pdam) == 0)  # how many had zero counts - only one sample FP108
ps_pdam <- prune_samples(sample_sums(ps_pdam) > 0, ps_pdam) #REMOVE FP108

sample_sums(ps_pdam)

#REmove samples with low read count - HB04 and SL51 never existed in the first place and FP54 and SL109 fastq files were corrupt. So 176 starting samples 

 #cutoff for samples is 10K read
#taking out: FP02 (5222), FP05 (3855), FP59 (9499), DL03 (9632), HB02 (2163), HB05 (1360), HB07 (1000), HB08 (5537), HB103 (162), HB106 (242), HB108 (8644), HB56 (3211), RC01 (6410), RC04 (29), RC07 (1384), RC101 (6338), RC102 (8220), RC108 (8609), RC53 (3708), SL08 (4088), SL10 (2), RF107 (7821) 
bad_samples <- c("FP02_R1.fastq.gz", "FP05_R1.fastq.gz",
                 "FP59_R1.fastq.gz", "DL03_R1.fastq.gz",
                 "HB02_R1.fastq.gz", "HB05_R1.fastq.gz", 
                 "HB07_R1.fastq.gz", "HB08_R1.fastq.gz",
                 "HB103_R1.fastq.gz", "HB106_R1.fastq.gz",
                 "HB108_R1.fastq.gz", "HB56_R1.fastq.gz",
                 "RC01_R1.fastq.gz", "RC04_R1.fastq.gz", 
                 "RC07_R1.fastq.gz", "RC101_R1.fastq.gz",
                 "RC102_R1.fastq.gz", "RC108_R1.fastq.gz",
                 "RC53_R1.fastq.gz", "SL08_R1.fastq.gz",
                 "SL10_R1.fastq.gz", "RF107_R1.fastq.gz")

ps_pdam  <- prune_samples(!(sample_names(ps_pdam ) %in% bad_samples), ps_pdam )
#n = 47/59 for Sept 2022, n = 55/58 for May 2024, n = 51/59 for Sept 2024

ps_pdam #153 samples 

```

```{r}
sample_data(ps_pdam)$Timepoint <- factor(sample_data(ps_pdam)$Timepoint,
                                         levels = c("Pre", "Bleach", "Post"),
                                         labels = timepoint_labels)

# Define and name timepoint colors
timepoint_labels <- c("Sept. 2022", "May 2024", "Sept. 2024")
timepoint_colors <- c("steelblue3", "violetred", "#1b9e77")
names(timepoint_colors) <- timepoint_labels


site_colors_micro <- c(
  "Harry's Bommie" = "#131142FF",
  "Fourth Point" = "dodgerblue2",
  "Deep Lagoon" = "steelblue1",
  "Shallow Lagoon" = "brown1",
  "Reef Crest" = "#C1000e",
  "Reef Flat" = "violetred")
```
#SEPT 2022 ANALYSIS 
```{r}

ps_t1 <- prune_samples(sample_data(ps_pdam)$Timepoint == "Sept. 2022", ps_pdam)


# Then prune taxa that have zero abundance across all remaining samples (this is leftover from the other timepoints)
ps_t1 <- prune_taxa(taxa_sums(ps_t1) > 0, ps_t1)

ps_t1 #2835 taxa, 47 samples 

```

#Rarefraction curve for read cutoff 
```{r}
otu <- as(otu_table(ps_t1), "matrix")

# Make sure samples are rows
if (taxa_are_rows(ps_t1)) {otu <- t(otu)}

# Rarefaction curves
rarecurve(otu, step = 1000, 
          sample = min(rowSums(otu)), 
          col = "darkblue", cex = 0.6, label = FALSE)

sample_sums(ps_t1)

```

#Alpha Diversity - Sept 2022
```{r}
set.seed(123)  # reproducibility
ps_t1_rare <- rarefy_even_depth(ps_t1, 
                                sample.size = 10000,   # cutoff depth
                                rngseed = 123,
                                replace = FALSE, 
                                verbose = TRUE)

```
```{r}
# Get richness estimates with rare sample metadata
richness <- estimate_richness(ps_t1_rare, measures = "Shannon")
richness$Origin <- sample_data(ps_t1_rare)$Origin
richness$Origin <- factor(richness$Origin, levels = c(
  "Harry's Bommie", 
  "Fourth Point", 
  "Deep Lagoon", 
  "Shallow Lagoon", 
  "Reef Crest", 
  "Reef Flat"))


shannondf <- summarySE(richness, measurevar= "Shannon", groupvars = c("Origin"), na.rm = TRUE)
shannondf


```

##stats
```{r}
#hist(pdamRichness$Shannon, main="Shannon index", xlab="")
pdamRichness_sept22 <- estimate_richness(ps_t1_rare, measures = "Shannon")
pdamRichness_sept22$Origin <- sample_data(ps_t1_rare)$Origin
head(pdamRichness_sept22)


# model + ANOVA (Type II) + Tukey
fit <- lm(Shannon ~ Origin, data = pdamRichness_sept22)
car::Anova(fit, type = 2)

tukey_shannon <- emmeans::emmeans(fit, pairwise ~ Origin, adjust = "tukey")
tukey_shannon


#NO SIGNIFICANCE 



```
##plot
```{r}
shannon_div <- ggplot(richness, aes(x = Origin, y = Shannon, fill = Origin)) +
  geom_boxplot(width = .5, outlier.shape = NA, alpha = 0.8, size = 1) +
    geom_jitter(width = 0.15, size = 4, alpha = 0.8, 
    shape = 21,stroke = 0.7) +
  scale_fill_manual(values = site_colors_micro, name = "Site",
    labels = c(
      "Harry's Bommie" = "HB",
      "Fourth Point" = "FP",
      "Deep Lagoon" = "DL",
      "Shallow Lagoon" = "SL",
      "Reef Crest" = "RC",
      "Reef Flat" = "RF")) +
  scale_x_discrete(labels = c(
    "Harry's Bommie" = "HB",
    "Fourth Point" = "FP",
    "Deep Lagoon" = "DL",
    "Shallow Lagoon" = "SL",
    "Reef Crest" = "RC",
    "Reef Flat" = "RF")) +
theme(
    axis.text.x  = element_blank(),
    axis.text.y = element_text(size = 18),
    axis.title.y = element_text(size = 18, face = "bold"),
    axis.title.x = element_blank(),
    panel.background = element_rect(fill = NA, color = "black"),
    legend.title = element_text(face = "bold", size = 16),
    legend.text = element_text(size = 12),
    #legend.position = "none"
    ) +
  labs(x = "Site", y = "Shannon Diversity\nIndex (H')")

shannon_div

```

#Beta Diversity 
```{r}

distNMDS_pst1 <- phyloseq::distance(ps_t1, method = "bray")
ordination_pst1 <- ordinate(ps_t1, method = "NMDS", distance = distNMDS_pst1)

ordination_pst1$stress

```
##Plot BD
```{r}
sample_data(ps_t1)$Origin <- factor(sample_data(ps_t1)$Origin,
                                    levels = c("Harry's Bommie",
                                               "Fourth Point",
                                               "Deep Lagoon",
                                               "Shallow Lagoon",
                                               "Reef Crest",
                                               "Reef Flat"))

p_t1_beta_NMDS <- plot_ordination(ps_t1, ordination_pst1, color = "Origin") +
  geom_point(size = 4) +
  stat_ellipse(aes(group = Origin, color = Origin), level = 0.7, size = 1) +
  scale_color_manual("Origin",
    values = c("Harry's Bommie"="#131142FF","Fourth Point"="dodgerblue2",
               "Deep Lagoon"="steelblue1","Shallow Lagoon"="brown1",
               "Reef Crest"="#C1000e","Reef Flat"="violetred")) +
  theme_classic() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme(
    panel.background = element_rect(fill = NA, color = "black"),
    #legend.position = "plot",
    legend.position = c(0.98, 0.98),
    legend.justification = c("right", "top"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    #legend.background = element_rect(fill = "white", color = NA)
  ) +
  annotate("text",
           x = -Inf, y = Inf,
           label = paste("Stress =", round(ordination_pst1$stress, 3)),
           hjust = -0.05, vjust = 1.1, size = 4)

p_t1_beta_NMDS
```
##Stats  
```{r}
meta.sept22 <- data.frame(sample_data(ps_t1))
meta.sept22 <- meta.sept22[match(rownames(as.matrix(distNMDS_pst1)), rownames(meta.sept22)), ]
#meta.sept22$Origin <- droplevels(factor(meta.sept22$Origin))

# PERMANOVA (effect of Origin on community composition)
perm_res <- adonis2(distNMDS_pst1 ~ Origin,
                    data = meta.sept22,
                    permutations = 9999,  # more perms = stabler p
                    by = "terms")
perm_res #is significant 


#pairwise 
pairwise.adonis(distNMDS_pst1, factors = sample_data(ps_t1)$Origin,
                                p.adjust.m = "BH")  #is signficant multiple interactions

#check betadispersion,
bd <- betadisper(distNMDS_pst1, meta.sept22$Origin)   # distances to group centroids
disp_res <- permutest(bd, permutations = 9999)
disp_res
#not significant at p = 0.3052


dispersion_df <- data.frame(
  DistanceToCentroid = bd$distances,
  Origin = meta.sept22$Origin)

dispersion_df$Origin <- factor(dispersion_df$Origin, levels = c(
  "Harry's Bommie", 
  "Fourth Point", 
  "Deep Lagoon", 
  "Shallow Lagoon", 
  "Reef Crest", 
  "Reef Flat"))


lm_disp <- lm(DistanceToCentroid ~ Origin, data = dispersion_df)
emmeans_disp <- emmeans(lm_disp, pairwise ~ Origin, adjust = "tukey")
emmeans_disp #pairwise not significant for beta dispersion for NMDS 

#BH (Benjamini–Hochberg) controls the false discovery rate (FDR)—the expected proportion of false positives among the pairs you call significant—while keeping more power than very conservative options (Bonferroni/Holm). It’s a good default for exploratory pairwise PERMANOVA tables.

```
##plot beta dispersion
```{r}
beta_dispersion_plot <- ggplot(dispersion_df, aes(x = Origin, y = DistanceToCentroid, fill = Origin)) +
    geom_boxplot(width = .5, outlier.shape = NA, alpha = 0.8, size = 1) +
    geom_jitter(width = 0.15, size = 4, alpha = 0.8, 
    shape = 21,stroke = 0.7) +
  scale_fill_manual(values = site_colors_micro,   name = "Site",labels = c(
      "Harry's Bommie" = "HB",
      "Fourth Point" = "FP",
      "Deep Lagoon" = "DL",
      "Shallow Lagoon" = "SL",
      "Reef Crest" = "RC",
      "Reef Flat" = "RF")) +
  scale_x_discrete(labels = c(
    "Harry's Bommie" = "HB",
    "Fourth Point" = "FP",
    "Deep Lagoon" = "DL",
    "Shallow Lagoon" = "SL",
    "Reef Crest" = "RC",
    "Reef Flat" = "RF")) +
  theme_classic() +
  theme(
    axis.text.x  = element_text(size =18),
    axis.text.y = element_text(size = 18),
    axis.title.y = element_text(size = 18, face = "bold"),
    axis.title.x = element_text(size = 18, face = "bold"),
    panel.background = element_rect(fill = NA, color = "black"),
    legend.title = element_text(face = "bold", size = 16),
    legend.text = element_text(size = 12),
    legend.position = "none") +
  labs(x = "Site", y = "Beta dispersion")

beta_dispersion_plot

```

#Relative Abundance
##Absolute abundance
```{r}
#absolute 
plot_bar(ps_t1, x="SampleID", fill="Phylum") + theme_classic() 
plot_bar(ps_t1, x="SampleID", fill="Family") + theme_classic()
plot_bar(ps_t1, x="SampleID", fill="Genus") + theme_classic() #too big to visualize

```
```{r}
#Relative abundance at different taxa levels 
pdam_fam <-tax_glom(ps_t1, "Family")
pdam_phy <-tax_glom(ps_t1, "Phylum")
pdam_gen <-tax_glom(ps_t1, "Genus")
pdam_order <-tax_glom(ps_t1, "Order")
```

```{r}
otu_table_ps <- otu_table(ps_t1)

# Transpose the OTU table so that samples are rows and taxa are columns
otu_table_ps <- t(otu_table_ps)

# Convert counts to relative abundance
otu_table_rel_abund <- otu_table_ps / rowSums(otu_table_ps)

head(otu_table_rel_abund)

otu_rel_abund_df <- as.data.frame(otu_table_rel_abund)
write.csv(otu_rel_abund_df, "C:/Users/marce/Desktop/Heron/CHOP_Heron_Jan2025/relative_abundance.csv")

```

##Phylum Level
```{r}

ps_s22_phy <- subset_samples(pdam_phy)
ps_s22_phy <- prune_taxa(taxa_sums(ps_s22_phy) > 0, ps_s22_phy)

pdam_phy_rel <- transform_sample_counts(ps_s22_phy, function(x) x/sum(x))
pdam_phy_rel <- prune_taxa(top20, pdam_phy_rel)
#pdam_fam_rel_abund <- ps_prune(pdam_fam_rel, min.abundance =0.03) #filter low abundant
pdam_phy_df <- psmelt(pdam_phy_rel)
```
```{r}
ggplot(pdam_phy_df, aes(x = Origin, y = Abundance, fill = Phylum)) + 
  theme_classic() + geom_bar(stat = "identity")

ggplot(pdam_phy_df, aes(x = SampleID, y = Abundance, fill = Phylum)) + 
  theme_classic() + geom_bar(stat = "identity") + facet_grid(~ Origin, space = "free", scales = "free") 

```
##Order level
```{r}

ps_s22_ord <- subset_samples(pdam_order)
ps_s22_ord <- prune_taxa(taxa_sums(ps_s22_ord) > 0, ps_s22_ord)

pdam_ord_rel <- transform_sample_counts(ps_s22_ord, function(x) x/sum(x))
pdam_ord_rel <- prune_taxa(top20, pdam_ord_rel)
pdam_ord_df <- psmelt(pdam_ord_rel)

ggplot(pdam_ord_df, aes(x = Origin, y = Abundance, fill = Order)) + 
  theme_classic() + geom_bar(stat = "identity")+ ggtitle("Relative Abundance: Top 20 Order")

ggplot(pdam_ord_df, aes(x = SampleID, y = Abundance, fill = Order)) + 
  theme_classic() + geom_bar(stat = "identity") + facet_grid(~ Origin, space = "free", scales = "free") +
  theme(axis.text.x = element_text(angle = 35, hjust=1)) + 
  labs(x = "Sample ID") + ggtitle("Relative Abundance:Order")

```
##Genus level
```{r}

#pdam_gen_t1 <-tax_glom(ps_t1, "Genus")

ps_s22 <- subset_samples(pdam_gen)
ps_s22 <- prune_taxa(taxa_sums(ps_s22) > 0, ps_s22) #removes 0 ASVs that were present in other timepoints 
ps_s22_rel <- transform_sample_counts(ps_s22, function(x) x / sum(x))

percentages_glom <- tax_glom(ps_s22_rel, taxrank = 'Genus')
#View(percentages_glom@tax_table@.Data)


df_s22 <- psmelt(percentages_glom)

df_s22$Genus <- as.factor(df_s22$Genus)
genus_colors <-colorRampPalette(brewer.pal(8,"Dark2")) (length(levels(df_s22$Genus)))
names(genus_colors) <- levels(df_s22$Genus)

df_s22$Genus <- as.character(df_s22$Genus) 

filtered_df_s22 <- df_s22 %>%
  dplyr::filter(Abundance >= 0.05)


#df_s22$Genus[df_s22$Abundance < 0.1] <- "Genus < 0.1% abund."
unique(filtered_df_s22$Genus)

abundance_sum<-summarySE(filtered_df_s22, measurevar='Abundance', groupvars=c('Origin','Genus', 'SampleID'), na.rm=TRUE, conf.interval = 0.95)
abundance_sum

df_s22$Origin <- factor(df_s22$Origin, levels = c(
  "Harry's Bommie", "Fourth Point", "Deep Lagoon",
  "Shallow Lagoon", "Reef Crest", "Reef Flat"))

# Plot
relative_plot_genus<- ggplot(abundance_sum, aes(x = SampleID, y = Abundance, fill = Genus)) + facet_wrap(~Origin, scales="free")+
  geom_bar(stat = "identity", color ="black", stroke =.25) +
  #scale_fill_manual(values = genus_colors) +
  theme_classic() +
  #scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  ) +
  labs(title = "Genus-Level Relative Abundance (Sept. 2022)",
       x = NULL, y = "Relative Abundance")

relative_plot_genus


```
##Family level
```{r}
ps_s22_fam <- subset_samples(pdam_fam)
ps_s22_fam <- prune_taxa(taxa_sums(ps_s22_fam) > 0, ps_s22_fam) #removes 0 ASVs that were present in other timepoints 

ps_s22_rel_fam <- transform_sample_counts(ps_s22_fam, function(x) x / sum(x))

percentages_glom_fam <- tax_glom(ps_s22_rel_fam, taxrank = 'Family')
#View(percentages_glom@tax_table@.Data)


df_s22_fam <- psmelt(percentages_glom_fam)
df_s22_fam$Family <- as.factor(df_s22_fam$Family)
df_s22_fam$Family <- as.character(df_s22_fam$Family) 

df_s22_fam$Origin <- factor(df_s22_fam$Origin, levels = c(
  "Harry's Bommie", "Fourth Point", "Deep Lagoon",
  "Shallow Lagoon", "Reef Crest", "Reef Flat"))

filtered_df_s22_fam <- df_s22_fam %>%
  dplyr::filter(Abundance >= 0.05)


#df_s22$Genus[df_s22$Abundance < 0.1] <- "Genus < 0.1% abund."
unique(filtered_df_s22_fam$Family)

abundance_sum_fam<-summarySE(filtered_df_s22_fam, measurevar='Abundance', groupvars=c('Origin','Family', 'SampleID'), na.rm=TRUE, conf.interval = 0.95)
abundance_sum_fam
write.csv(abundance_sum_fam, "rel_abundance_filtered.csv", row.names = FALSE)

abundance_sum_fam <- abundance_sum_fam %>%
  mutate(Family = factor(Family)) %>%
  droplevels()

families_present <- levels(abundance_sum_fam$Family)
n_fam <- length(families_present)

# Build a Paired palette sized to n_fam
base_pal <- brewer.pal(12, "Paired")
if (n_fam <= 12) {
  pal <- brewer.pal(max(3, n_fam), "Paired")[seq_len(n_fam)]
} else {
  pal <- colorRampPalette(base_pal)(n_fam)  # extend beyond 12
}
names(pal) <- families_present

relative_plot_fam <- ggplot(abundance_sum_fam, aes(x = SampleID, y = Abundance, fill = Family)) +
  facet_wrap(~Origin, scales = "free_x", nrow = 1, ncol = 6) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.25) +
  scale_fill_manual(values = pal,
                    breaks = families_present,
                    limits = families_present,
                    drop   = TRUE) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 16), axis.title.y = element_text(size = 16, face = "bold"), axis.text.y = element_text(size=16),legend.title = element_text(size = 16, face = "bold"),
        legend.position = "right", legend.text = element_text(size = 12),strip.text = element_text(size = 16, face = "bold")) +
  labs(x = NULL, y = "Family-level\nrelative abundance")

relative_plot_fam

```

#Microbial PCA
##PCA w/distance to centroids
```{r}
ps_clr <- microbiome::transform(ps_s22_fam, "clr")

# Extract CLR OTU table
otu_clr <- otu_table(ps_clr)
if (taxa_are_rows(ps_clr)) otu_clr <- t(otu_clr)

pca_res <- prcomp(otu_clr, center = TRUE, scale. = FALSE)

# Build PCA dataframe
pca_df <- as.data.frame(pca_res$x) %>%
  rownames_to_column("SampleID")

#extract metadata
metadata <- as(sample_data(ps_s22_fam), "data.frame")

# Ensure SampleID is a column (overwrite with rownames if needed)
metadata$SampleID <- rownames(metadata)

# Merge PCA results with metadata
pca_df <- left_join(pca_df, metadata, by = "SampleID")

var_exp <- (pca_res$sdev^2) / sum(pca_res$sdev^2)
x_lab <- sprintf("PC1 (%.1f%% explained var.)", 100 * var_exp[1])
y_lab <- sprintf("PC2 (%.1f%% explained var.)", 100 * var_exp[2])

# PCA plot with percent on axes
p_pca <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Origin)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA of CLR-Transformed Microbiome Data",
       x = x_lab, y = y_lab) +
  theme(plot.title = element_text(hjust = 0.5))

p_pca


var_exp <- (pca_res$sdev^2) / sum(pca_res$sdev^2)
data.frame(PC = seq_along(var_exp),
           prop = var_exp,
           cum  = cumsum(var_exp))[1:10, ]
```

```{r}
loadings_df <- as.data.frame(pca_res$rotation[, 1:2])  # PC1 and PC2 only
loadings_df$Taxon <- rownames(loadings_df)

loadings_df$Magnitude <- sqrt(loadings_df$PC1^2 + loadings_df$PC2^2)

loadings_df <- tibble::as_tibble(loadings_df)

top_taxa <- dplyr::arrange(loadings_df, desc(Magnitude)) %>%
  dplyr::slice(1:20)


tax_df <- tax_table(ps_t1) %>%
  as.data.frame() %>%
  rownames_to_column("Taxon")

# Merge taxonomy with top_taxa
top_taxa_annotated <- left_join(top_taxa, tax_df, by = "Taxon")

top_taxa_annotated <- top_taxa_annotated %>% #fix labels 
  mutate(Label = ifelse(!is.na(Family), Family,
                        ifelse(!is.na(Family), Family, Taxon)))


scale_factor <- 100 

p_pca_biplot <- p_pca +
  geom_segment(data = top_taxa_annotated,
               aes(x = 0, y = 0,
                   xend = PC1 * scale_factor,
                   yend = PC2 * scale_factor),
               arrow = arrow(length = unit(0.2, "cm")),color = "black",inherit.aes = FALSE) +
  geom_text(data = top_taxa_annotated,
            aes(x = PC1 * scale_factor * 1.06,
                y = PC2 * scale_factor * 1.06,
                label = Label),size = 4,hjust = 0.5,inherit.aes = FALSE)

p_pca_biplot

```
```{r}

centroids_df <- pca_df %>%
  group_by(Origin) %>%
  dplyr::summarize(PC1 = mean(PC1),
            PC2 = mean(PC2),
            .groups = "drop")

desired_order <- c("Harry's Bommie", "Fourth Point", "Deep Lagoon", "Shallow Lagoon", "Reef Crest", "Reef Flat")

centroids_df$Origin <- factor(centroids_df$Origin, levels = desired_order)

centroids_df <- centroids_df %>%
  arrange(factor(Origin, levels = desired_order))

label_offset <- 0.08

lab_df <- top_taxa_annotated %>%
  mutate(
    len = sqrt(PC1^2 + PC2^2),
    x_tip = PC1 * scale_factor,
    y_tip = PC2 * scale_factor,
    nudge_x = (PC1 / len) * label_offset,
    nudge_y = (PC2 / len) * label_offset)

p_pca_centroids <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Origin)) +
  geom_point(size = 4, alpha = 0.8) + stat_ellipse(aes(color = Origin),level = 0.8, geom = "path", linetype = "solid", size = 1) +
  geom_segment(data = top_taxa_annotated,
               aes(x = 0, y = 0,
                   xend = PC1 * scale_factor,
                   yend = PC2 * scale_factor),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "black", inherit.aes = FALSE) +
  geom_text_repel(
    data = lab_df,
    aes(x = x_tip, y = y_tip, label = Label),
    nudge_x = lab_df$nudge_x,
    nudge_y = lab_df$nudge_y,
    size = 5,
    box.padding = 0.35,
    point.padding = 0.25,
    min.segment.length = 0,
    segment.size = 0.3,
    seed = 42,
    max.overlaps = Inf, color = "black") +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.8, color = "darkgray") + 
  geom_vline(xintercept = 0, linetype = 2, alpha = 0.8, color = "darkgray") +
  scale_color_manual(values = site_colors_micro) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.text  = element_text(size = 18),
    panel.background = element_rect(fill = NA, color = 'black'),
    #legend.position = c(0.02, 0.98),
    legend.position = "none",
    #legend.justification = c("left", "top"),
    plot.margin = margin(10, 20, 10, 20)) +
  coord_cartesian(clip = "off") +
  labs(x = x_lab, y = y_lab)

p_pca_centroids
```

##stats PCA 
```{r}

adon_one <- adonis2(dist_clr ~ Origin, data = pca_df,
                    permutations = 9999, by = "margin")
adon_one
#significant effect of origin

rownames(pca_df) <- pca_df$SampleID
pairwise_results_micro <- pairwise.adonis(
  x = as.matrix(dist_clr),
  factors = pca_df$Origin,
  perm = 9999,
  p.adjust.m = "BH")

pairwise_results_micro #not significant but that;s okay

```

##environmental inset - PCA 
```{r}

env_vars 

env_summary <- phys %>%
  dplyr::select(Origin, all_of(env_vars)) %>%
  distinct(Origin, .keep_all = TRUE)

env_summary <- env_summary %>%
  mutate(
    Origin = dplyr::recode(
      Origin,
      "HB" = "Harry's Bommie",
      "FP" = "Fourth Point",
      "DL" = "Deep Lagoon",
      "SL" = "Shallow Lagoon",
      "RC" = "Reef Crest",
      "RF" = "Reef Flat"))

#check that names match microbiome metadata
unique(env_summary$Origin)
unique(pca_df$Origin)

centroids_env <- pca_df %>%
  group_by(Origin) %>%
  dplyr::summarize(
    PC1 = mean(PC1, na.rm = TRUE),
    PC2 = mean(PC2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(env_summary, by = "Origin")

# Check for missing environmental data after join
centroids_env %>%
  filter(if_any(all_of(env_vars), is.na)) %>%
  dplyr::select(Origin)


micro_envfit <- envfit(
  centroids_env[, c("PC1", "PC2")],
  centroids_env[, env_vars],
  na.rm = TRUE,
  permutations = 9999)

micro_envfit

env_scores_micro <- as.data.frame(micro_envfit$vectors$arrows * sqrt(micro_envfit$vectors$r))

env_scores_micro$var <- rownames(env_scores_micro)

env_scores_micro$label <- dplyr::recode(
  env_scores_micro$var,
  mean_o2 = "DO",
  mean_salinity = "Salinity",
  mean_turbidity = "Turbidity",
  mean_depth = "Depth",
  mean.seasonal.amp = "Seasonal~amp.",
  mean.daily.amp = "Daily~amp.",
  DLI = "PAR~dose",
  Fluor = "Fluor.",
  O2_amp = "DO~amp.",)
```

```{r}
env_scores_micro <- env_scores_micro %>%
  mutate(
    arrow_x = PC1 * 1.4,
    arrow_y = PC2 * 1.4,
    label_x = PC1 * 1.6,
    label_y = PC2 * 1.6)


env_inset_micro <- ggplot() +
  geom_hline(yintercept = 0,linetype = 2,alpha = 0.5,color = "gray50") +
  geom_vline(xintercept = 0,linetype = 2,alpha = 0.5,color = "gray50") +
  geom_segment(
    data = env_scores_micro,
    aes(x = 0, y = 0, xend = arrow_x, yend = arrow_y),
    arrow = arrow(length = unit(0.2, "lines")),
    color = "black",
    alpha = 0.9) +
  geom_text_repel(data = env_scores_micro,aes(x = label_x, y = label_y, label = label),size = 5,color = "black",box.padding = 0.25, point.padding = 0.15,force = 5,force_pull = 0.2,max.overlaps = Inf,segment.color = NA,seed = 123,parse = TRUE) +
  theme_classic() +
  theme(
    plot.margin = margin(0, 0, 0, 0),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = "black"),
    panel.border = element_rect(color = "black", fill = NA),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none")

env_inset_micro

```

##beta dispersion
```{r}
dist_clr <- dist(otu_clr, method = "euclidean")

# Grouping variable for centroid distances
group_var <- sample_data(ps_s22_fam)$Origin 
# Run beta dispersion analysis
bd_micro <- betadisper(dist_clr, group_var)

distances_df_micro <- data.frame(
  DistanceToCentroidMicro = bd_micro$distances,
  Origin = group_var)

bd_perm <- permutest(bd_micro, permutations = 9999)  # global test
bd_perm
#beta dispersion is not significant 


TukeyHSD(bd_micro)

```
```{r}

distances_df_micro$Origin <- factor(distances_df_micro$Origin, levels = c(
  "Harry's Bommie", 
  "Fourth Point", 
  "Deep Lagoon", 
  "Shallow Lagoon", 
  "Reef Crest", 
  "Reef Flat"))


p_dist <- ggplot(distances_df_micro, aes(x = Origin, y = DistanceToCentroidMicro, fill = Origin)) +
  geom_boxplot(width = .5, outlier.shape = NA, alpha = 0.8, size = 1) +
    geom_jitter(width = 0.15, size = 4, alpha = 0.8, 
    shape = 21,stroke = 0.7) +
  scale_fill_manual(values = site_colors_micro,   name = "Site",labels = c(
      "Harry's Bommie" = "HB",
      "Fourth Point" = "FP",
      "Deep Lagoon" = "DL",
      "Shallow Lagoon" = "SL",
      "Reef Crest" = "RC",
      "Reef Flat" = "RF")) +
  scale_x_discrete(labels = c(
    "Harry's Bommie" = "HB",
    "Fourth Point" = "FP",
    "Deep Lagoon" = "DL",
    "Shallow Lagoon" = "SL",
    "Reef Crest" = "RC",
    "Reef Flat" = "RF")) +
  scale_y_continuous(n.breaks = 8) +
  theme_classic() +
  theme(
    axis.text.x  = element_text(size =18),
    axis.text.y = element_text(size = 18),
    axis.title.y = element_text(size = 18, face = "bold"),
    axis.title.x = element_text(size = 18, face = "bold"),
    panel.background = element_rect(fill = NA, color = "black"),
    legend.title = element_text(face = "bold", size = 16),
    legend.text = element_text(size = 12),
    #legend.position = "none"
  ) + labs(x = "Site", y = "Distance to\ncentroid")
p_dist
```
#FIGURE 5
```{r}

legend_shared <- get_legend(
  shannon_div +
    theme(
      legend.position = "right",
      legend.title = element_text(size = 16, face = "bold"),
      legend.text = element_text(size = 14)))

shannon_div <- shannon_div + theme(legend.position = "none")
pdist   <- p_dist + theme(legend.position = "none")

left_bottom <- plot_grid(
  p_pca_centroids,
  env_inset_micro,
  ncol = 2,
  rel_widths = c(1.8, 0.9),
  labels = c("B", "C"),
  label_size = 18,
  label_fontface = "bold",
  label_x = c(0, -0.05),
  label_y = c(1, 1),
  hjust = -0.5,
  vjust = 1.5,
  align = "h"
)

right_plots <- plot_grid(
  shannon_div,
  pdist,
  ncol = 1,
  labels = c("D", "E"),
  label_size = 18,
  label_fontface = "bold",
  rel_heights = c(1, 1),
  align = "v"
)

right_bottom <- plot_grid(
  right_plots,
  legend_shared,
  ncol = 2,
  rel_widths = c(1, 0.35)
)

bottom_row <- plot_grid(
  left_bottom,
  right_bottom,
  ncol = 2,
  rel_widths = c(3.8, 1.2),
  align = "h"
)

micro_full_fig <- plot_grid(
  relative_plot_fam,
  bottom_row,
  ncol = 1,
  rel_heights = c(1.3, 2.2),
  labels = c("A", ""),
  label_size = 18,
  label_fontface = "bold"
)

micro_full_fig

```

#Spearman Correlations
##overall phys x microbe
```{r}
phys22 <- read.csv("C:/Users/marce/Desktop/Heron/CHOP_Heron_Jan2025/PBR_pre_data.csv")
rownames(phys22) <- phys22$SampleID
head(phys22)

sample_data_22 <- sample_data(ps_s22_rel_fam)  #using family level relative abundance 

# Replace row names with the SampleID
sample_names(ps_s22_rel_fam) <- sample_data_22$SampleID

ps_s22_clr_fam <- microbiome::transform(ps_s22_rel_fam, "clr")

# Extract CLR-transformed abundance table as data frame
otu_fam <- as.data.frame(otu_table(ps_s22_clr_fam))

# Verify that samples are rows and families are columns
if (taxa_are_rows(ps_s22_clr_fam)) {
  otu_fam <- t(otu_fam)
}
head(otu_fam)

# Optional: check distribution / sanity of transformed values
summary(otu_fam[, 1:5])


```
```{r}
merged_data <- merge(otu_fam, phys22, by = "row.names", all = TRUE)
rownames(merged_data) <- merged_data$Row.names
merged_data <- merged_data[, -1]  # Remove Row.names column

# View the merged data
head(merged_data)
numeric_data <- merged_data[sapply(merged_data, is.numeric)]

# Check the structure of the numeric data to ensure it's all numeric
str(numeric_data)


taxa_sums <- colSums(otu_fam)

# Get the top 30
top30_taxa <- names(sort(taxa_sums, decreasing = TRUE))[1:30]

# Subset the OTU table to these top 30 taxa
otu_top30 <- otu_fam[, top30_taxa]

merged_top30 <- merge(otu_top30, phys22, by = "row.names", all = FALSE)
rownames(merged_top30) <- merged_top30$Row.names
merged_top30 <- merged_top30[, -1]  # Remove Row.names column
numeric_top30 <- merged_top30[sapply(merged_top30, is.numeric)]

```
```{r}

correlation_matrix <- cor(numeric_data, method = "spearman", use = "pairwise.complete.obs")
```
```{r}
melted_corr <- reshape2::melt(correlation_matrix)
```
```{r}

taxa_sums <- colSums(otu_fam)
top30_taxa <- names(sort(taxa_sums, decreasing = TRUE))[1:30]

# Subset the OTU table to these top 30 taxa
otu_top30 <- otu_fam[, top30_taxa]

merged_top30 <- merge(otu_top30, phys22, by = "row.names", all = FALSE)
rownames(merged_top30) <- merged_top30$Row.names
merged_top30 <- merged_top30[, -1]  # Remove Row.names column

numeric_top30 <- merged_top30[sapply(merged_top30, is.numeric)]

# Create correlation matrix
cor_matrix_top30 <- cor(numeric_top30, method = "spearman", use = "pairwise.complete.obs")

melted_corr_top30 <- reshape2::melt(cor_matrix_top30)

# Plot heatmap
# ggplot(data = melted_corr_top30, aes(Var1, Var2, fill = value)) +
#   geom_tile() +
#   scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
#   labs(title = "Correlation Matrix: Top 30 Taxa + Physiology", x = "Variables", y = "Variables")


```

```{r}
dist_mat <- as.dist(1 - cor_matrix_top30)

# Hierarchical clustering
hc <- hclust(dist_mat)

# Get the order of variables
ordered_vars <- hc$labels[hc$order]

# Apply that to the melted dataframe
melted_corr_top30$Var1 <- factor(melted_corr_top30$Var1, levels = ordered_vars)
melted_corr_top30$Var2 <- factor(melted_corr_top30$Var2, levels = ordered_vars)

# Plot
# ggplot(data = melted_corr_top30, aes(Var1, Var2, fill = value)) +
#   geom_tile() +
#   scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
#   labs(title = "Clustered Correlation Matrix", x = "Variables", y = "Variables")

```

```{r}
#Create ASV -> Family mapping
tax_df <- as.data.frame(tax_table(ps_s22_rel_fam))
tax_df$ASV <- rownames(tax_df)
tax_df$Family[is.na(tax_df$Family)] <- "Unclassified_Family"
tax_df$Family_Labeled <- make.unique(tax_df$Family)
asv_to_family <- setNames(tax_df$Family_Labeled, tax_df$ASV)

```

##unclustered
```{r}
phys_vars <- colnames(phys22)[sapply(phys22, is.numeric)]
microbe_vars <- colnames(otu_top30)[1:30]  

microbe_family_names <- asv_to_family[microbe_vars]

# Subset the correlation matrix
cor_subset <- cor_matrix_top30[phys_vars, microbe_vars]


cor_subset <- cor_matrix_top30[phys_vars, microbe_vars]
colnames(cor_subset) <- microbe_family_names
# Melt the matrix for ggplot
melted_subset <- reshape2::melt(cor_subset)

# Plot with phys on y and microbe on x
ggplot(data = melted_subset, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Physiology vs Top 30 Microbiome Correlations", x = "Microbiome Taxa", y = "Physiological Variables")
```

##clustered
```{r}
phys_vars <- colnames(phys22)[sapply(phys22, is.numeric)]
microbe_vars <- colnames(otu_top30)[1:30]
microbe_family_names <- asv_to_family[microbe_vars]

common_samples <- intersect(rownames(phys22), rownames(otu_top30))

merged_df <- cbind(
  phys22[common_samples, phys_vars],
  otu_top30[common_samples, microbe_vars]
)
rcorr_res <- rcorr(as.matrix(merged_df), type = "spearman")

cor_subset <- rcorr_res$r[phys_vars, microbe_vars, drop = FALSE]
pval_subset <- rcorr_res$P[phys_vars, microbe_vars, drop = FALSE]

colnames(cor_subset) <- microbe_family_names[colnames(cor_subset)]
colnames(pval_subset) <- microbe_family_names[colnames(pval_subset)]

row_order <- hclust(as.dist(1 - cor(t(cor_subset))))$order
col_order <- hclust(as.dist(1 - cor(cor_subset)))$order

cor_clustered <- cor_subset[row_order, col_order, drop = FALSE]
pval_clustered <- pval_subset[row_order, col_order, drop = FALSE]

melted_cor <- reshape2::melt(cor_clustered)
melted_pval <- reshape2::melt(pval_clustered)

melted_cor$stars <- cut(melted_pval$value,
                        breaks = c(-Inf, 0.001, 0.01, 0.05, 1),
                        labels = c("***", "**", "*", ""),
                        right = FALSE)

pretty_phys_labels <- c(
  "pre_protein" = "Host protein",
  "pre_AFDW" = "Total biomass",
  "pre_chlor" = "Chlorophyll a",
  "pre_symb" = "Symbiont density",
  "pre_poc" = "Pocilloporin",
  "pre_MG" = "Mycosporine-glycine",
  "Fv.Fm" = "Fv/Fm",
  "ED50" = "Thermal tolerance (ED50)",
  "pre_den" = "Skeletal density")

 combined_site_corr_plot <- ggplot(data = melted_cor, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "gray90") +
  geom_text(aes(label = stars), size = 10, vjust = 0.5, hjust = 0.5) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, name = "Correlation") +
  scale_y_discrete(labels = pretty_phys_labels) +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 16, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 16), axis.title.x = element_text(size=16), axis.title.y = element_text(size=16), legend.title = element_text(size = 16),legend.text  = element_text(size = 16)) +
  labs(x = "Microbial family", y = "Physiological variable")
 
combined_site_corr_plot 

```

##Individual Sites - Unclustered 
###FP
```{r}
site <- "FP"

# Subset phys and micro data
site_phys <- phys22[phys22$Origin == site, ]
site_micro <- otu_top30[rownames(otu_top30) %in% rownames(site_phys), microbe_vars, drop = FALSE]

# Merge data
merged_site <- merge(site_phys, site_micro, by = "row.names")
rownames(merged_site) <- merged_site$Row.names
merged_site <- merged_site[, !(names(merged_site) %in% "Row.names")]

# Keep numeric only
merged_numeric <- merged_site[, sapply(merged_site, is.numeric)]

# Correlation matrix and p-values
rcorr_res <- rcorr(as.matrix(merged_numeric), type = "spearman")
cor_site <- rcorr_res$r
pval_site <- rcorr_res$P

# Subset to phys × microbes
available_phys <- intersect(phys_vars, rownames(cor_site))
available_microbes <- intersect(microbe_vars, colnames(cor_site))

cor_subset_site <- cor_site[available_phys, available_microbes, drop = FALSE]
pval_subset_site <- pval_site[available_phys, available_microbes, drop = FALSE]

# Rename microbial columns to family names
colnames(cor_subset_site) <- microbe_family_names[available_microbes]
colnames(pval_subset_site) <- microbe_family_names[available_microbes]

# Drop one physiological variable
cor_subset_site <- cor_subset_site[rownames(cor_subset_site) != "ED50", , drop = FALSE]
pval_subset_site <- pval_subset_site[rownames(pval_subset_site) != "ED50", , drop = FALSE]

# Keep only physiological rows with at least one significant correlation - comment out if you want the full matrix
sig_rows <- rownames(pval_subset_site)[apply(pval_subset_site < 0.05, 1, any, na.rm = TRUE)]

cor_subset_site <- cor_subset_site[sig_rows, , drop = FALSE]
pval_subset_site <- pval_subset_site[sig_rows, , drop = FALSE]

# Melt matrices
melted_site <- reshape2::melt(cor_subset_site)
melted_pval <- reshape2::melt(pval_subset_site)
melted_site$pval <- melted_pval$value

# Add significance stars
melted_site$stars <- cut(
  melted_site$pval,
  breaks = c(-Inf, 0.001, 0.01, 0.05, 1),
  labels = c("***", "**", "*", ""),
  right = FALSE
)

# Custom Y-axis order
custom_order <- c(
  "pre_den",      # Skeletal Density
  "pre_AFDW",    # Total Biomass
  "pre_protein", # Host Protein
  "pre_poc",     # Pocilloporin
  "pre_MG",     # Mycosporine-glycine
  "pre_symb",    # Symbiont Density
  "pre_chlor",   # Chlorophyll a
  "Fv.Fm"      # Photochemical efficiency
  
  
)

# Keep only rows still present after filtering for sig rows - comment out if you want the full matrix
custom_order <- custom_order[custom_order %in% rownames(cor_subset_site)]

melted_site$Var1 <- factor(melted_site$Var1, levels = rev(custom_order))
melted_site$Var2 <- factor(melted_site$Var2, levels = colnames(cor_subset_site))  # lock x-axis too

# Plot
FP_unclust <- ggplot(data = melted_site, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "gray90") +
  geom_text(aes(label = stars), size = 10, fontface = "bold",color = "black", na.rm = TRUE) +
  scale_fill_gradient2(
    low = "blue", high = "red", mid = "white",
    midpoint = 0, limits = c(-1, 1)
  ) +
  scale_y_discrete(labels = function(x) parse(text = pretty_phys_labels[x])) +
  coord_equal() +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 16),
    plot.title = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16, face = "bold")
  ) +
  labs(title = "Fourth Point", fill = "Correlation")

FP_unclust

```
###RF
```{r}
site <- "RF"

# Subset phys and micro data
site_phys <- phys22[phys22$Origin == site, ]
site_micro <- otu_top30[rownames(otu_top30) %in% rownames(site_phys), microbe_vars, drop = FALSE]

# Merge data
merged_site <- merge(site_phys, site_micro, by = "row.names")
rownames(merged_site) <- merged_site$Row.names
merged_site <- merged_site[, !(names(merged_site) %in% "Row.names")]

# Keep numeric only
merged_numeric <- merged_site[, sapply(merged_site, is.numeric)]

# Correlation matrix and p-values
rcorr_res <- rcorr(as.matrix(merged_numeric), type = "spearman")
cor_site <- rcorr_res$r
pval_site <- rcorr_res$P

# Subset to phys × microbes
available_phys <- intersect(phys_vars, rownames(cor_site))
available_microbes <- intersect(microbe_vars, colnames(cor_site))

cor_subset_site <- cor_site[available_phys, available_microbes, drop = FALSE]
pval_subset_site <- pval_site[available_phys, available_microbes, drop = FALSE]

# Rename microbial columns to family names
colnames(cor_subset_site) <- microbe_family_names[available_microbes]
colnames(pval_subset_site) <- microbe_family_names[available_microbes]

# Drop one physiological variable
cor_subset_site <- cor_subset_site[rownames(cor_subset_site) != "ED50", , drop = FALSE]
pval_subset_site <- pval_subset_site[rownames(pval_subset_site) != "ED50", , drop = FALSE]

## Keep only physiological rows with at least one significant correlation - comment out if you want the full matrix
sig_rows <- rownames(pval_subset_site)[apply(pval_subset_site < 0.05, 1, any, na.rm = TRUE)]

cor_subset_site <- cor_subset_site[sig_rows, , drop = FALSE]
pval_subset_site <- pval_subset_site[sig_rows, , drop = FALSE]

# Melt matrices
melted_site <- reshape2::melt(cor_subset_site)
melted_pval <- reshape2::melt(pval_subset_site)
melted_site$pval <- melted_pval$value

# Add significance stars
melted_site$stars <- cut(
  melted_site$pval,
  breaks = c(-Inf, 0.001, 0.01, 0.05, 1),
  labels = c("***", "**", "*", ""),
  right = FALSE
)

# Custom Y-axis order
custom_order <- c(
  "pre_den",      # Skeletal Density
  "pre_AFDW",    # Total Biomass
  "pre_protein", # Host Protein
  "pre_poc",     # Pocilloporin
  "pre_MG",     # Mycosporine-glycine
  "pre_symb",    # Symbiont Density
  "pre_chlor",   # Chlorophyll a
  "Fv.Fm"      # Photochemical efficiency
  
  
)

# Keep only rows still present after filtering for sig rows - comment out if you want the full matrix
custom_order <- custom_order[custom_order %in% rownames(cor_subset_site)]

melted_site$Var1 <- factor(melted_site$Var1, levels = rev(custom_order))
melted_site$Var2 <- factor(melted_site$Var2, levels = colnames(cor_subset_site))  # lock x-axis too

# Plot
RF_unclust <- ggplot(data = melted_site, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "gray90") +
  geom_text(aes(label = stars), size = 10, fontface = "bold", color = "black", na.rm = TRUE) +
  scale_fill_gradient2(
    low = "blue", high = "red", mid = "white",
    midpoint = 0, limits = c(-1, 1)
  ) +
  scale_y_discrete(labels = function(x) parse(text = pretty_phys_labels[x])) +
  theme_minimal() +
  coord_equal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_blank(),
    plot.title = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16, face = "bold")
  ) +
  labs(
    title = "Reef Flat",
    x = "Microbial Family",
    fill = "Correlation"
  )

RF_unclust

```
###HB
```{r}
site <- "HB"

# Subset phys and micro data
site_phys <- phys22[phys22$Origin == site, ]
site_micro <- otu_top30[rownames(otu_top30) %in% rownames(site_phys), microbe_vars, drop = FALSE]

# Merge data
merged_site <- merge(site_phys, site_micro, by = "row.names")
rownames(merged_site) <- merged_site$Row.names
merged_site <- merged_site[, !(names(merged_site) %in% "Row.names")]

# Keep numeric only
merged_numeric <- merged_site[, sapply(merged_site, is.numeric)]

# Correlation matrix and p-values
rcorr_res <- rcorr(as.matrix(merged_numeric), type = "spearman")
cor_site <- rcorr_res$r
pval_site <- rcorr_res$P

# Subset to phys × microbes
available_phys <- intersect(phys_vars, rownames(cor_site))
available_microbes <- intersect(microbe_vars, colnames(cor_site))

cor_subset_site <- cor_site[available_phys, available_microbes, drop = FALSE]
pval_subset_site <- pval_site[available_phys, available_microbes, drop = FALSE]

# Rename microbial columns to family names
colnames(cor_subset_site) <- microbe_family_names[available_microbes]
colnames(pval_subset_site) <- microbe_family_names[available_microbes]

# Drop one physiological variable
cor_subset_site <- cor_subset_site[rownames(cor_subset_site) != "ED50", , drop = FALSE]
pval_subset_site <- pval_subset_site[rownames(pval_subset_site) != "ED50", , drop = FALSE]

# Keep only physiological rows with at least one significant correlation - comment out if you want the full matrix
sig_rows <- rownames(pval_subset_site)[apply(pval_subset_site < 0.05, 1, any, na.rm = TRUE)]

cor_subset_site <- cor_subset_site[sig_rows, , drop = FALSE]
pval_subset_site <- pval_subset_site[sig_rows, , drop = FALSE]

# Melt matrices
melted_site <- reshape2::melt(cor_subset_site)
melted_pval <- reshape2::melt(pval_subset_site)
melted_site$pval <- melted_pval$value

# Add significance stars
melted_site$stars <- cut(
  melted_site$pval,
  breaks = c(-Inf, 0.001, 0.01, 0.05, 1),
  labels = c("***", "**", "*", ""),
  right = FALSE
)

# Custom Y-axis order
custom_order <- c(
  "pre_den",      # Skeletal Density
  "pre_AFDW",    # Total Biomass
  "pre_protein", # Host Protein
  "pre_poc",     # Pocilloporin
  "pre_MG",     # Mycosporine-glycine
  "pre_symb",    # Symbiont Density
  "pre_chlor",   # Chlorophyll a
  "Fv.Fm"      # Photochemical efficiency
  
  
)

# Keep only rows still present after filtering for sig rows - comment out if you want the full matrix
custom_order <- custom_order[custom_order %in% rownames(cor_subset_site)]

melted_site$Var1 <- factor(melted_site$Var1, levels = rev(custom_order))
melted_site$Var2 <- factor(melted_site$Var2, levels = colnames(cor_subset_site))  # lock x-axis too

# Plot
HB_unclust <- ggplot(data = melted_site, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "gray90") +
  geom_text(aes(label = stars), size = 10, fontface = "bold", color = "black", na.rm = TRUE) +
  scale_fill_gradient2(
    low = "blue", high = "red", mid = "white",
    midpoint = 0, limits = c(-1, 1)
  ) +
  coord_equal() +
  scale_y_discrete(labels = function(x) parse(text = pretty_phys_labels[x])) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 16),
    plot.title = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16, face = "bold")
  ) +
  labs(title = "Harry's Bommie", fill = "Correlation")

HB_unclust

```
###SL
```{r}

site <- "SL"

# Subset phys and micro data
site_phys <- phys22[phys22$Origin == site, ]
site_micro <- otu_top30[rownames(otu_top30) %in% rownames(site_phys), microbe_vars, drop = FALSE]

# Merge data
merged_site <- merge(site_phys, site_micro, by = "row.names")
rownames(merged_site) <- merged_site$Row.names
merged_site <- merged_site[, !(names(merged_site) %in% "Row.names")]

# Keep numeric only
merged_numeric <- merged_site[, sapply(merged_site, is.numeric)]

# Correlation matrix and p-values
rcorr_res <- rcorr(as.matrix(merged_numeric), type = "spearman")
cor_site <- rcorr_res$r
pval_site <- rcorr_res$P

# Subset to phys × microbes
available_phys <- intersect(phys_vars, rownames(cor_site))
available_microbes <- intersect(microbe_vars, colnames(cor_site))

cor_subset_site <- cor_site[available_phys, available_microbes, drop = FALSE]
pval_subset_site <- pval_site[available_phys, available_microbes, drop = FALSE]

# Rename microbial columns to family names
colnames(cor_subset_site) <- microbe_family_names[available_microbes]
colnames(pval_subset_site) <- microbe_family_names[available_microbes]

# Drop one physiological variable
cor_subset_site <- cor_subset_site[rownames(cor_subset_site) != "ED50", , drop = FALSE]
pval_subset_site <- pval_subset_site[rownames(pval_subset_site) != "ED50", , drop = FALSE]

# Keep only physiological rows with at least one significant correlation - comment out if you want the full matrix
sig_rows <- rownames(pval_subset_site)[apply(pval_subset_site < 0.05, 1, any, na.rm = TRUE)]

cor_subset_site <- cor_subset_site[sig_rows, , drop = FALSE]
pval_subset_site <- pval_subset_site[sig_rows, , drop = FALSE]

# Melt matrices
melted_site <- reshape2::melt(cor_subset_site)
melted_pval <- reshape2::melt(pval_subset_site)
melted_site$pval <- melted_pval$value

# Add significance stars
melted_site$stars <- cut(
  melted_site$pval,
  breaks = c(-Inf, 0.001, 0.01, 0.05, 1),
  labels = c("***", "**", "*", ""),
  right = FALSE
)

# Custom Y-axis order
custom_order <- c(
  "pre_den",      # Skeletal Density
  "pre_AFDW",    # Total Biomass
  "pre_protein", # Host Protein
  "pre_poc",     # Pocilloporin
  "pre_MG",     # Mycosporine-glycine
  "pre_symb",    # Symbiont Density
  "pre_chlor",   # Chlorophyll a
  "Fv.Fm"      # Photochemical efficiency
  
  
)

# Keep only rows still present after filtering for sig rows - comment out if you want the full matrix
custom_order <- custom_order[custom_order %in% rownames(cor_subset_site)]

melted_site$Var1 <- factor(melted_site$Var1, levels = rev(custom_order))
melted_site$Var2 <- factor(melted_site$Var2, levels = colnames(cor_subset_site))  # lock x-axis too

# Plot
SL_unclust <- ggplot(data = melted_site, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "gray90") +
  geom_text(aes(label = stars), size = 10, fontface = "bold", color = "black", na.rm = TRUE) +
  scale_fill_gradient2(
    low = "blue", high = "red", mid = "white",
    midpoint = 0, limits = c(-1, 1)
  ) +
  coord_equal() +
  scale_y_discrete(labels = function(x) parse(text = pretty_phys_labels[x])) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 16),
    plot.title = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16, face = "bold")
  ) +
  labs(title = "Shallow Lagoon", fill = "Correlation")

SL_unclust


```
###RC
```{r}
site <- "RC"

# Subset phys and micro data
site_phys <- phys22[phys22$Origin == site, ]
site_micro <- otu_top30[rownames(otu_top30) %in% rownames(site_phys), microbe_vars, drop = FALSE]

# Merge data
merged_site <- merge(site_phys, site_micro, by = "row.names")
rownames(merged_site) <- merged_site$Row.names
merged_site <- merged_site[, !(names(merged_site) %in% "Row.names")]

# Keep numeric only
merged_numeric <- merged_site[, sapply(merged_site, is.numeric)]

# Correlation matrix and p-values
rcorr_res <- rcorr(as.matrix(merged_numeric), type = "spearman")
cor_site <- rcorr_res$r
pval_site <- rcorr_res$P

# Subset to phys × microbes
available_phys <- intersect(phys_vars, rownames(cor_site))
available_microbes <- intersect(microbe_vars, colnames(cor_site))

cor_subset_site <- cor_site[available_phys, available_microbes, drop = FALSE]
pval_subset_site <- pval_site[available_phys, available_microbes, drop = FALSE]

# Rename microbial columns to family names
colnames(cor_subset_site) <- microbe_family_names[available_microbes]
colnames(pval_subset_site) <- microbe_family_names[available_microbes]

# Drop one physiological variable
cor_subset_site <- cor_subset_site[rownames(cor_subset_site) != "ED50", , drop = FALSE]
pval_subset_site <- pval_subset_site[rownames(pval_subset_site) != "ED50", , drop = FALSE]

# Keep only physiological rows with at least one significant correlation - comment out if you want the full matrix
sig_rows <- rownames(pval_subset_site)[apply(pval_subset_site < 0.05, 1, any, na.rm = TRUE)]

cor_subset_site <- cor_subset_site[sig_rows, , drop = FALSE]
pval_subset_site <- pval_subset_site[sig_rows, , drop = FALSE]

# Melt matrices
melted_site <- reshape2::melt(cor_subset_site)
melted_pval <- reshape2::melt(pval_subset_site)
melted_site$pval <- melted_pval$value

# Add significance stars
melted_site$stars <- cut(
  melted_site$pval,
  breaks = c(-Inf, 0.001, 0.01, 0.05, 1),
  labels = c("***", "**", "*", ""),
  right = FALSE
)

# Custom Y-axis order
custom_order <- c(
  "pre_den",      # Skeletal Density
  "pre_AFDW",    # Total Biomass
  "pre_protein", # Host Protein
  "pre_poc",     # Pocilloporin
  "pre_MG",     # Mycosporine-glycine
  "pre_symb",    # Symbiont Density
  "pre_chlor",   # Chlorophyll a
  "Fv.Fm"      # Photochemical efficiency
  
  
)

# Keep only rows still present after filtering for sig rows - comment out if you want the full matrix
custom_order <- custom_order[custom_order %in% rownames(cor_subset_site)]

melted_site$Var1 <- factor(melted_site$Var1, levels = rev(custom_order))
melted_site$Var2 <- factor(melted_site$Var2, levels = colnames(cor_subset_site))  # lock x-axis too

# Plot
RC_unclust <- ggplot(data = melted_site, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "gray90") +
  geom_text(aes(label = stars), size = 10, fontface = "bold", color = "black", na.rm = TRUE) +
  scale_fill_gradient2(
    low = "blue", high = "red", mid = "white",
    midpoint = 0, limits = c(-1, 1)
  ) +
  coord_equal() +
  scale_y_discrete(labels = function(x) parse(text = pretty_phys_labels[x])) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 16),
    plot.title = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16, face = "bold")
  ) +
  labs(title = "Reef Crest", fill = "Correlation")

RC_unclust

```
###DL
```{r}
site <- "DL"

# Subset phys and micro data
site_phys <- phys22[phys22$Origin == site, ]
site_micro <- otu_top30[rownames(otu_top30) %in% rownames(site_phys), microbe_vars, drop = FALSE]

# Merge data
merged_site <- merge(site_phys, site_micro, by = "row.names")
rownames(merged_site) <- merged_site$Row.names
merged_site <- merged_site[, !(names(merged_site) %in% "Row.names")]

# Keep numeric only
merged_numeric <- merged_site[, sapply(merged_site, is.numeric)]

# Correlation matrix and p-values
rcorr_res <- rcorr(as.matrix(merged_numeric), type = "spearman")
cor_site <- rcorr_res$r
pval_site <- rcorr_res$P

# Subset to phys × microbes
available_phys <- intersect(phys_vars, rownames(cor_site))
available_microbes <- intersect(microbe_vars, colnames(cor_site))

cor_subset_site <- cor_site[available_phys, available_microbes, drop = FALSE]
pval_subset_site <- pval_site[available_phys, available_microbes, drop = FALSE]

# Rename microbial columns to family names
colnames(cor_subset_site) <- microbe_family_names[available_microbes]
colnames(pval_subset_site) <- microbe_family_names[available_microbes]

# Drop one physiological variable
cor_subset_site <- cor_subset_site[rownames(cor_subset_site) != "ED50", , drop = FALSE]
pval_subset_site <- pval_subset_site[rownames(pval_subset_site) != "ED50", , drop = FALSE]

# Keep only physiological rows with at least one significant correlation - comment out if you want the full matrix
sig_rows <- rownames(pval_subset_site)[apply(pval_subset_site < 0.05, 1, any, na.rm = TRUE)]

cor_subset_site <- cor_subset_site[sig_rows, , drop = FALSE]
pval_subset_site <- pval_subset_site[sig_rows, , drop = FALSE]

# Melt matrices
melted_site <- reshape2::melt(cor_subset_site)
melted_pval <- reshape2::melt(pval_subset_site)
melted_site$pval <- melted_pval$value

# Add significance stars
melted_site$stars <- cut(
  melted_site$pval,
  breaks = c(-Inf, 0.001, 0.01, 0.05, 1),
  labels = c("***", "**", "*", ""),
  right = FALSE
)

# Custom Y-axis order
custom_order <- c(
  "pre_den",      # Skeletal Density
  "pre_AFDW",    # Total Biomass
  "pre_protein", # Host Protein
  "pre_poc",     # Pocilloporin
  "pre_MG",     # Mycosporine-glycine
  "pre_symb",    # Symbiont Density
  "pre_chlor",   # Chlorophyll a
  "Fv.Fm")      # Photochemical efficiency

# Keep only rows still present after filtering for sig rows - comment out if you want the full matrix
custom_order <- custom_order[custom_order %in% rownames(cor_subset_site)]

melted_site$Var1 <- factor(melted_site$Var1, levels = rev(custom_order))
melted_site$Var2 <- factor(melted_site$Var2, levels = colnames(cor_subset_site))  # lock x-axis too

# Plot
DL_unclust <- ggplot(data = melted_site, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "gray90") +
  geom_text(aes(label = stars), size = 10, fontface = "bold", hjust = 0.5, vjust = 0.5,color = "black", na.rm = TRUE) +
  scale_fill_gradient2(
    low = "blue", high = "red", mid = "white",
    midpoint = 0, limits = c(-1, 1)
  ) +
  coord_equal() +
  scale_y_discrete(labels = function(x) parse(text = pretty_phys_labels[x])) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 16),
    plot.title = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16, face = "bold")
  ) +
  labs(title = "Deep Lagoon", fill = "Correlation")

DL_unclust

```
#FIGURE 6
```{r}
library('ggpubr')

layout <- "
A
B
C
D
E
F
"

combined_plot_7_V2 <- (
  HB_unclust + FP_unclust + DL_unclust + SL_unclust + RC_unclust + RF_unclust) +
  plot_layout(design = layout, guides = "collect") + plot_annotation(tag_levels = 'A') &
  theme(
    plot.tag = element_text(size = 16, face = "bold")
  )
combined_plot_7_V2

```