---
title: "Reef-scale environmental mosaics shape coral physiological and microbial signatures
  of resilience"
author: "Marcelina Martynek"
date: "2026-05-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("ggplot2")
library("tidyverse")
library('csv')
library('mgcv')
library('patchwork')
library('gridExtra')
library('cowplot')
library('emmeans')
library('reshape2')
library('ggrepel')
library('ggfortify') 
library('Rmisc')
library(beeswarm)
library(dplyr)
library(lubridate)
library(grid)
library(vegan)
library(pairwiseAdonis)

phys<-read.csv("Pdam comparative physiology CURRENT.csv", strip.white=T)
phys$Origin<-as.factor(phys$Origin)
phys$Coral.ID<-as.factor(phys$Coral.ID)
phys$mean.daily.amp<-as.numeric(phys$mean.daily.amp)
phys


site_colors_ID <- c(
  "RF" = "violetred",  
  "RC" = "#C1000e",  
  "SL" = "brown1", 
  "DL" = "steelblue1", 
  "FP" = "dodgerblue2", 
  "HB" = "#131142FF")
```

#Physiology - all 
#Protein
###Normality
```{r}
hist(phys$Protein)
```

```{r}
protein.lm<- lm(Protein~Origin, data=phys) #Gaussian
par(mfrow=c(2,3))
plot(protein.lm, ask=FALSE, which=1:6)
```
###Stats
```{r}
protein.lme <- lme(Protein~Origin, random = ~1|Coral.ID, data=phys, na.action=na.exclude)
car::Anova(protein.lme, type=2)
tukey<- emmeans(protein.lme, list(pairwise ~ Origin), adjust = "tukey")
tukey
```
###Figure
```{r}
phys_sum_prot <- summarySE(
  data = phys,
  measurevar = "Protein",
  groupvars = c("Origin")
)

phys_sum_prot$Origin <- factor(phys_sum_prot$Origin, levels = c("HB", "FP", "DL", "SL", "RC", "RF"))
prot_line_plot <- ggplot(phys_sum_prot, aes(x = Origin, y = Protein)) +
  #geom_line(aes(group = 1), size = 1, color = "gray40") +  
  geom_point(aes(color = Origin), size = 5) +              
  geom_errorbar(aes(ymin = Protein - se, ymax = Protein + se, color = Origin), width = 0.2) +
  scale_y_continuous(expression("       Host-soluble\nprotein content"~(mg~'\U2219'~cm^{-2})), limits = c(0, 1.5)) +
  scale_color_manual(values = site_colors_ID) +
  theme_classic() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 16),
    legend.position = "none",
    panel.background = element_rect(fill = NA, color = 'black')
  )

prot_line_plot

```

#Symbiont density
###Normality
```{r}
hist(phys$Sym.density)
hist(phys$Regate)
```

```{r}
sym.lm<- lm(Regate~Origin, data=phys) #Gaussian
par(mfrow=c(2,3))
plot(sym.lm, ask=FALSE, which=1:6)
```
###Stats
```{r}
sym.lme <- lme(Regate~Origin, random = ~1|Coral.ID, data=phys, na.action=na.exclude)
car::Anova(sym.lme, type=2)
tukey<- emmeans(sym.lme, list(pairwise ~ Origin), adjust = "tukey")
tukey
```
###Figure
```{r}
phys_sum_sym <- summarySE(
  data = phys,
  measurevar = "Regate",
  groupvars = c("Origin"), 
  na.rm = TRUE)


phys_sum_sym$Origin <- factor(phys_sum_sym$Origin, levels = c("HB", "FP", "DL", "SL", "RC", "RF"))
sym_line_plot <- ggplot(phys_sum_sym, aes(x = Origin, y = Regate)) +
  #geom_line(aes(group = 1), size = 1, color = "gray40") +  
  geom_point(aes(color = Origin), size = 5) +         
  geom_errorbar(aes(ymin = Regate - se, ymax = Regate + se, color = Origin), width = 0.2) +
  scale_y_continuous(expression(Symbiont~density~(10^{6}~cells~'\U2219'~cm^{-2})), limits = c(0, 0.8)) +
  scale_color_manual(values = site_colors_ID) +
  theme_classic() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 16),
    legend.position = "none",
    panel.background = element_rect(fill = NA, color = 'black')
  )

sym_line_plot
```

#CaCO3 density
###Normality
```{r}
hist(phys$CaCO3.Density..g.ml.)
```

```{r}
density.lm<- lm(CaCO3.Density..g.ml.~Origin, data=phys) #Gaussian
par(mfrow=c(2,3))
plot(density.lm, ask=FALSE, which=1:6)
```
###Stats
```{r}
density.lme <- lme(CaCO3.Density..g.ml.~Origin, random = ~1|Coral.ID, data=phys, na.action=na.exclude)
car::Anova(density.lme, type=2)
tukey<- emmeans(density.lme, list(pairwise ~ Origin), adjust = "tukey")
tukey
```
#Figure
```{r}
phys_sum_den <- summarySE(
  data = phys,
  measurevar = "CaCO3.Density..g.ml.",
  groupvars = c("Origin"), 
  na.rm = TRUE)

phys_sum_den$Origin <- factor(phys_sum_den$Origin, levels = c("HB", "FP", "DL", "SL", "RC", "RF"))
den_line_plot <- ggplot(phys_sum_den, aes(x = Origin, y = CaCO3.Density..g.ml.)) +
  #geom_line(aes(group = 1), size = 1, color = "gray40") + 
  geom_point(aes(color = Origin), size = 5) +            
  geom_errorbar(aes(ymin = CaCO3.Density..g.ml. - se, ymax = CaCO3.Density..g.ml. + se, color = Origin), width = 0.2) +
  scale_y_continuous(expression(CaCO[3]~density~(g~'\U2219'~mL^{-1})), limits = c(1, 3), breaks = seq(1, 3, by = 1)) +
  scale_color_manual(values = site_colors_ID) +
  theme_classic() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 16),
    legend.position = "none",
    panel.background = element_rect(fill = NA, color = 'black'))

den_line_plot
```

#Chlorophyll
###Normality
```{r}
hist(phys$Chlor)
```

```{r}
chlor.lm<- lm(Chlor~Origin, data=phys) #Gaussian
par(mfrow=c(2,3))
plot(chlor.lm, ask=FALSE, which=1:6)
```
###Stats
```{r}
chlor.lme <- lme(Chlor~Origin, random = ~1|Coral.ID, data=phys, na.action=na.exclude)
car::Anova(chlor.lme, type=2)
tukey<- emmeans(chlor.lme, list(pairwise ~ Origin), adjust = "tukey")
tukey
```
###Figure
```{r}
phys_sum_chlor <- summarySE(
  data = phys,
  measurevar = "Chlor",
  groupvars = c("Origin"), 
  na.rm = TRUE
)

phys_sum_chlor$Origin <- factor(phys_sum_chlor$Origin, levels = c("HB", "FP", "DL", "SL", "RC", "RF"))
chlor_line_plot <- ggplot(phys_sum_chlor, aes(x = Origin, y = Chlor)) +
  #geom_line(aes(group = 1), size = 1, color = "gray40") +
  geom_point(aes(color = Origin), size = 5) +           
  geom_errorbar(aes(ymin = Chlor - se, ymax = Chlor + se, color = Origin), width = 0.2) +
  scale_y_continuous(expression(Chlorophyll~a~(mu~g~'\U2219'~cm^{-2})), limits = c(0, 3), breaks = seq(0, 3, by = .5)) +
  scale_color_manual(values = site_colors_ID) +
  theme_classic() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 16),
    legend.position = "none",
    panel.background = element_rect(fill = NA, color = 'black')
  )

chlor_line_plot

```

#AFDW
###Normality
```{r}
hist(phys$AFDW..mg.cm.2.)


#removing one outlier 
phys_withoutAFDWoutlier <- phys

phys_withoutAFDWoutlier[57, 11] <- NA  #SL07

```

```{r}
AFDW.lm<- lm(AFDW..mg.cm.2.~Origin, data=phys_withoutAFDWoutlier) #Gaussian
par(mfrow=c(2,3))
plot(AFDW.lm, ask=FALSE, which=1:6)
```
###Stats
```{r}
AFDW.lme <- lme(AFDW..mg.cm.2.~Origin, random = ~1|Coral.ID, data=phys_withoutAFDWoutlier, na.action=na.exclude)
car::Anova(AFDW.lme, type=2)
tukey<- emmeans(AFDW.lme, list(pairwise ~ Origin), adjust = "tukey")
tukey
```
###Figure
```{r}
phys_sum_AFDW <- summarySE(
  data = phys_withoutAFDWoutlier,
  measurevar = "AFDW..mg.cm.2.",
  groupvars = c("Origin"), 
  na.rm = TRUE
)

phys_sum_AFDW$Origin <- factor(phys_sum_AFDW$Origin, levels = c("HB", "FP", "DL", "SL", "RC", "RF"))
AFDW_line_plot <- ggplot(phys_sum_AFDW, aes(x = Origin, y = AFDW..mg.cm.2.)) +
  #geom_line(aes(group = 1), size = 1, color = "gray40") +  
  geom_point(aes(color = Origin), size = 5) +             
  geom_errorbar(aes(ymin = AFDW..mg.cm.2. - se, ymax = AFDW..mg.cm.2. + se, color = Origin), width = 0.2) +
  scale_y_continuous(expression(Total~biomass~(mg~'\U2219'~cm^{-2})), limits = c(0, 2.5)) +
  scale_color_manual(values = site_colors_ID) +
  theme_classic() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 16),
    legend.position = "none",
    panel.background = element_rect(fill = NA, color = 'black'))

AFDW_line_plot

```
#MMAs
#Removing outliers in MAAs 
```{r}

#Testing showed that Sample SL07 is way off the charts
phys_withoutMAAoutliers <- phys

phys_withoutMAAoutliers[57, 3] <- NA  # Mycrosporine..glycine
phys_withoutMAAoutliers[57, 4] <- NA  # Palythine
phys_withoutMAAoutliers[57, 5] <- NA  # Asterina
phys_withoutMAAoutliers[57, 6] <- NA  # Shinorine

# Check for NA values in the specific locations
phys_withoutMAAoutliers[57, c(3, 4, 5, 6)]

##after removing SL07, RC04 proved to be an outlier as well. Removing 
phys_withoutMAAoutliers[34, 3] <- NA  # Mycrosporine..glycine
phys_withoutMAAoutliers[34, 4] <- NA  # Palythine
phys_withoutMAAoutliers[34, 5] <- NA  # Asterina
phys_withoutMAAoutliers[34, 6] <- NA  # Shinorine

##after removing RC04, FP10 proved to be an outlier as well. Removing 
phys_withoutMAAoutliers[20, 3] <- NA  # Mycrosporine..glycine
phys_withoutMAAoutliers[20, 4] <- NA  # Palythine
phys_withoutMAAoutliers[20, 5] <- NA  # Asterina
phys_withoutMAAoutliers[20, 6] <- NA  # Shinorine

```

#Mycrosporine glycine
###Normality
```{r}
hist(phys_withoutMAAoutliers$Mycrosporine..glycine)
```

```{r}
maa1.lm<- lm(Mycrosporine..glycine~Origin, data=phys_withoutMAAoutliers) #Gaussian
par(mfrow=c(2,3))
plot(maa1.lm, ask=FALSE, which=1:6)
```
###Stats
```{r}
maa1.lme <- lme(Mycrosporine..glycine~Origin, random = ~1|Coral.ID, data=phys_withoutMAAoutliers, na.action=na.exclude)
car::Anova(maa1.lme, type=2)
tukey<- emmeans(maa1.lme, list(pairwise ~ Origin), adjust = "tukey")
tukey
```
#Palythine
###Normality
```{r}
hist(phys$Palythine)
```

```{r}
pal.lm<- lm(Palythine~Origin, data=phys_withoutMAAoutliers) #Gaussian
par(mfrow=c(2,3))
plot(pal.lm, ask=FALSE, which=1:6)
```
###Stats
```{r}
pal.lme <- lme(Palythine~Origin, random = ~1|Coral.ID, data=phys_withoutMAAoutliers, na.action=na.exclude)
car::Anova(pal.lme, type=2)
tukey<- emmeans(pal.lme, list(pairwise ~ Origin), adjust = "tukey")
tukey
```
#Asterina
###Normality
```{r}
hist(phys$Asterina)
```

```{r}
ast.lm<- lm(Asterina~Origin, data=phys_withoutMAAoutliers) #Gaussian
par(mfrow=c(2,3))
plot(ast.lm, ask=FALSE, which=1:6)
```
###Stats
```{r}
ast.lme <- lme(Asterina~Origin, random = ~1|Coral.ID, data=phys_withoutMAAoutliers, na.action=na.exclude)
car::Anova(ast.lme, type=2)
tukey<- emmeans(ast.lme, list(pairwise ~ Origin), adjust = "tukey")
tukey
```
#Shinorine
###Normality
```{r}
hist(phys$Shinorine)
```

```{r}
shin.lm<- lm(Shinorine~Origin, data=phys_withoutMAAoutliers) #Gaussian
par(mfrow=c(2,3))
plot(shin.lm, ask=FALSE, which=1:6)
```
###Stats
```{r}
shin.lme <- lme(Shinorine~Origin, random = ~1|Coral.ID, data=phys_withoutMAAoutliers, na.action=na.exclude)
car::Anova(shin.lme, type=2)
tukey<- emmeans(shin.lme, list(pairwise ~ Origin), adjust = "tukey")
tukey
```

#All MMA Plot 
```{r}
phys_long_MAA <- phys_withoutMAAoutliers %>%
  pivot_longer(
    cols = c(Mycrosporine..glycine, Shinorine, Palythine, Asterina),
    names_to = "MAA_type",
    values_to = "MAA_conc"
  )

phys_sum_MAA <- summarySE(
  data = phys_long_MAA,
  measurevar = "MAA_conc",
  groupvars = c("Origin", "MAA_type"),
  na.rm = TRUE
)
```

```{r}

phys_sum_MAA$Origin <- factor(phys_sum_MAA$Origin, levels = c("HB", "FP", "DL", "SL", "RC", "RF"))
dodge <- position_dodge(width = 0.5)

MAA_line_plot <- ggplot(phys_sum_MAA, aes(x = Origin, y = MAA_conc, group = MAA_type)) +
  geom_point(aes(color = Origin, shape = MAA_type), 
             size = 5, position = dodge) +
  geom_errorbar(aes(ymin = MAA_conc - se, ymax = MAA_conc + se, color = Origin), 
                width = 0.2, position = dodge) +
  scale_y_continuous(expression(MAA~concentration~(m~g~'\U2219'~cm^{-2})), limits = c(0, 12.5), breaks = seq(0, 12.5, by = 2.5)) +
  scale_color_manual(values = site_colors_ID) +
  scale_shape_manual(
    values = c(16, 17, 15, 18),
    name = "MAA Type",
    labels = c("Mycosporine-glycine", "Shinorine", "Palythine", "Asterina")
  ) +
  guides(color = "none") +
  theme_classic() +
theme(
  legend.position = "none",
  panel.background = element_rect(fill = NA, color = 'black'), 
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16)
)

MAA_line_plot

```

#Pocilloporin
###Normality
```{r}
hist(phys$Pocilloporin)

#One outlier --> 34 (RC04) 
phys_withoutPOCoutliers <- phys

phys_withoutPOCoutliers[34, 7] <- NA  # Pocilloporin

# Check for NA values in the specific locations
phys_withoutPOCoutliers[34, 7]

#After removing the first outlier, SL07 & FP10 proved to be an outlier as well. Removing
phys_withoutPOCoutliers[57, 7] <- NA
phys_withoutPOCoutliers[20, 7] <- NA


```

```{r}
poc.lm<- lm(Pocilloporin~Origin, data=phys_withoutPOCoutliers) #Gaussian
par(mfrow=c(2,3))
plot(poc.lm, ask=FALSE, which=1:6)
```
###Stats
```{r}
poc.lme <- lme(Pocilloporin~Origin, random = ~1|Coral.ID, data=phys_withoutPOCoutliers, na.action=na.exclude)
car::Anova(poc.lme, type=2)
tukey<- emmeans(poc.lme, list(pairwise ~ Origin), adjust = "tukey")
tukey
```
###Figure
```{r}
phys_sum_poc <- summarySE(
  data = phys_withoutPOCoutliers,
  measurevar = "Pocilloporin",
  groupvars = c("Origin"), 
  na.rm = TRUE
)


phys_sum_poc$Origin <- factor(phys_sum_poc$Origin, levels = c("HB", "FP", "DL", "SL", "RC", "RF"))
poc_line_plot <- ggplot(phys_sum_poc, aes(x = Origin, y = Pocilloporin)) +
  #geom_line(aes(group = 1), size = 1, color = "gray40") +  
  geom_point(aes(color = Origin), size = 5) +           
  geom_errorbar(aes(ymin = Pocilloporin - se, ymax = Pocilloporin + se, color = Origin), width = 0.2) +
  scale_y_continuous(expression(Pocilloporin~(m~g~'\U2219'~cm^{-2})), limits =c(0, 3)) +
  scale_color_manual(values = site_colors_ID) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    legend.position = "none",
    panel.background = element_rect(fill = NA, color = 'black')
  )

poc_line_plot

```
#ED50
###Stats
```{r}
phys_sum_ED50 <- data.frame(
  Origin = c("HB", "FP", "DL", "SL", "RC", "RF"),
  ED50 = c(35.371, 35.099, 36.067, 36.810, 35.888, 36.080),
  se = c(0.224, 0.238, 0.474, 0.928, 0.520, 0.459))

```
###Figure
```{r}
phys_sum_ED50$Origin <- factor(phys_sum_ED50$Origin, levels = c("HB", "FP", "DL", "SL", "RC", "RF"))
ED50_line_plot <- ggplot(phys_sum_ED50, aes(x = Origin, y = ED50)) +
  #geom_line(aes(group = 1), size = 1, color = "gray40") + 
  geom_point(aes(color = Origin), size = 5) +            
  geom_errorbar(aes(ymin = ED50 - se, ymax = ED50 + se, color = Origin), width = 0.2) +
  scale_y_continuous(expression("Thermal tolerance (ED50)"), breaks = seq(34, 39, by = 0.5)) +
  scale_color_manual(values = site_colors_ID) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    legend.position = "none",
    panel.background = element_rect(fill = NA, color = 'black')
  )

ED50_line_plot
```

#Fv/Fm
```{r}
hist(phys$FvFm)
```
###Normality
```{r}
fvfm.lm<- lm(FvFm~Origin, data=phys) #Gaussian
par(mfrow=c(2,3))
plot(fvfm.lm, ask=FALSE, which=1:6)
```
###Stats
```{r}
fvfm.lme <- lme(FvFm~Origin, random = ~1|Coral.ID, data=phys, na.action=na.exclude)
car::Anova(fvfm.lme, type=2)
tukey<- emmeans(fvfm.lme, list(pairwise ~ Origin), adjust = "tukey")
tukey
```
###Figure
```{r}
phys_sum_FvFm <- summarySE(
  data = phys,
  measurevar = "FvFm",
  groupvars = c("Origin"), 
  na.rm = TRUE)

phys_sum_FvFm$Origin <- factor(phys_sum_FvFm$Origin, levels = c("HB", "FP", "DL", "SL", "RC", "RF"))
FvFm_line_plot <- ggplot(phys_sum_FvFm, aes(x = Origin, y = FvFm)) +
  #geom_line(aes(group = 1), size = 1, color = "gray40") +  
  geom_point(aes(color = Origin), size = 5) +             
  geom_errorbar(aes(ymin = FvFm - se, ymax = FvFm + se, color = Origin), width = 0.2) +
  scale_y_continuous(expression("Photochemical yield (Fv/Fm)"), limits = c(0.5, 0.7)) +
  scale_color_manual(values = site_colors_ID) +
  theme_classic() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 16),
    legend.position = "none",
    panel.background = element_rect(fill = NA, color = 'black'))

FvFm_line_plot

```

#SUPP FIG 2 - PHYSIOLOGY 
```{r}
legend_plot <- ggplot(phys_sum_prot, aes(x = Origin, y = Protein, color = Origin)) + #get common legend 
  geom_point() +
  scale_color_manual(values = site_colors_ID) +
  theme_void() +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.title = element_blank(),
    legend.text = element_text(size = 9),               
    legend.key.width = unit(1.0, "cm"),                 
    legend.key.height = unit(0.4, "cm"),
    legend.spacing.x = unit(0.3, 'cm'))

legend <- get_legend(legend_plot)

main_grid <- (
  prot_line_plot + AFDW_line_plot + den_line_plot +
  sym_line_plot + chlor_line_plot + FvFm_line_plot +
  MAA_line_plot + poc_line_plot + ED50_line_plot) +
  plot_layout(ncol = 3) +
  plot_annotation(tag_levels = 'A') &
    theme(plot.tag = element_text(face = 'bold', size =16))

combined_plot_phys <- plot_grid(
  #legend,
  main_grid,
  ncol = 1,
  rel_heights = c(0.1, 1))


combined_plot_phys
```

#Multivariate Physiology - PCA
```{r}
colnames(phys)
```

```{r}

#sum(is.na(phys[,(3:13)]))  # Count NA values in the selected columns
phys_clean <- phys[-c(34, 57, 20),] # Remove rows 34, 57, 20 (outliers) SL07, RC04, FP10

phys.rda<- rda(phys_clean[, c(3:8, 10:13, 24:25)], na.rm=TRUE, scale=TRUE)
```

```{r}
biplot(phys.rda, scaling=1, main='Scaling')
```

```{r}
pca.sites.scores<- as.data.frame(scores(phys.rda, display='sites'))
pca.sites.scores<- data.frame(pca.sites.scores, phys_clean)
pca.species.scores<- as.data.frame(scores(phys.rda, display='species'))
pca.species.scores$Species<- rownames(pca.species.scores)
pca.species.scores
```
```{r}
include_list<- c("Mycrosporine..glycine", "Asterina", "Shinorine", "Palythine", "Pocilloporin", "CaCO3.Density..g.ml.", "Protein", "AFDW..mg.cm.2.", "Sym.density", "Chlor", "ED50", "FvFm") #List of vectors to inclue on the below plot
pca.species.scores<-pca.species.scores[include_list,]
```

```{r} 
eig<-eigenvals(phys.rda)
eig
```


###stats - PCA 
```{r}

phys.adonis<-adonis2(phys_clean[, c(3:8, 10:13, 24:25)] ~Origin, data=phys_clean, permutations = 9999, na.rm=T)
phys.adonis


pairwise_phys <- pairwise.adonis2(phys_clean[, c(3:8, 10:13, 24:25)] ~Origin, data=phys_clean, permutations = 9999, na.rm=T)

pair_tab <- as.data.frame(pairwise_phys$pairs)  
pair_tab

```

###collapse MAAs 
```{r}

pca.sites.scores$Origin <- factor(pca.sites.scores$Origin, levels = c("RF", "RC", "SL", "DL", "FP", "HB"))

pca.variables <- pca.species.scores[1:12, ]  # force 12 rows

pca.variables$Species <- c(
  "Mycosporine glycine", "Asterina", "Shinorine", "Palythine", 
  "Pocilloporin", "Skeletal density", "Host protein", 
  "Total biomass", "Symbiont density", "Chlorophyll-a", 
  "ED50", "Fv/Fm")

MAA_group <- pca.variables %>%
  filter(Species %in% c("Mycosporine glycine", "Asterina", "Shinorine", "Palythine")) %>%
  dplyr::summarize(
    PC1 = mean(PC1),
    PC2 = mean(PC2),
    Species = "MMAs",
    .groups = "drop")

non_MAA_group <- pca.variables %>%
  filter(!Species %in% c("Mycosporine glycine", "Asterina", "Shinorine", "Palythine"))

pca.variables <- bind_rows(MAA_group, non_MAA_group) %>%
  mutate(
      label = case_when(
      Species == "Fv/Fm" ~ "F[v]/F[m]",
      TRUE ~ paste0("'", Species, "'")))
```
###plot PCA
```{r}

g <- ggplot() +
  geom_point(data = pca.sites.scores, aes(x = PC1, y = PC2, color = Origin), size = 5, alpha = 0.8) +
  stat_ellipse(data = pca.sites.scores, aes(x = PC1, y = PC2, group = Origin, color = Origin), 
               type = "norm", level = 0.5, size = 1, linetype = "solid") +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.8, color = "darkgray") + 
  geom_vline(xintercept = 0, linetype = 2, alpha = 0.8, color = "darkgray") +
  geom_segment(data = pca.variables, aes(x = 0, y = 0, xend = PC1 *1.5, yend = PC2 *1.5), 
               arrow = arrow(length = unit(.3, 'lines')), 
               color = 'black', linetype = 'solid', alpha = 0.9) +
  geom_text_repel(data = pca.variables, aes(x = PC1 *1.6, y = PC2 *1.6, label = label), color = "black", parse = TRUE, size = 8, 
  nudge_x = 0, nudge_y = 0) +
  scale_color_manual(name = "Site",values = site_colors_ID) +
  scale_x_continuous(limits = c(-1.5, 2.5),
    paste0(names(eig[1]), sprintf(' (%0.1f%% explained var.)', 100 * eig[1] / sum(eig)))) +
  scale_y_continuous(
    limits = c(-1.5, 2.5),
    paste0(names(eig[2]), sprintf(' (%0.1f%% explained var.)', 100 * eig[2] / sum(eig)))) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.position = c(0.98, 0.98),
  legend.justification = c(1, 1),
  legend.title = element_text(size = 18, face = "bold"),
  legend.text = element_text(size = 16),
  legend.key.size = unit(0.8, "cm"),
    panel.background = element_rect(fill = NA, color = 'black'),
    strip.text = element_text(size = 12, face = "italic")) 

g
```

###distance to centroid: Phys PCA
```{r}
pca.sites.scores <- as.data.frame(scores(phys.rda, display = "sites", choices = 1:12))
pca.sites.scores$Origin <- phys_clean$Origin
pca.sites.scores$Origin <- factor(pca.sites.scores$Origin, levels = c("HB", "FP", "DL", "SL", "RC", "RF"))


centroids <- pca.sites.scores %>%
  group_by(Origin) %>%
  dplyr::summarize(across(PC1:PC12, mean, na.rm = TRUE), .groups = "drop")

dist_pca <- pca.sites.scores %>%
  left_join(centroids, by = "Origin", suffix = c("", "_centroid")) %>%
  rowwise() %>%
  dplyr::mutate(dist_to_centroid = sqrt(sum((c_across(PC1:PC12) - c_across(PC1_centroid:PC12_centroid))^2))) %>%
  ungroup()


anova_model <- lm(dist_to_centroid ~ Origin, data = dist_pca)
car::Anova(anova_model, type = 2)


tukey_dist_phys<- emmeans(anova_model, list(pairwise ~ Origin), adjust = "tukey")
tukey_dist_phys

```

```{r}

phys_dist <- ggplot(dist_pca, aes(x = Origin, y = dist_to_centroid, fill = Origin)) +
  geom_boxplot(width = .5, outlier.shape = NA, alpha = 0.8, size = 1) +
    geom_jitter(width = 0.15, size = 4, alpha = 0.8, 
    shape = 21,stroke = 0.7) +
  scale_fill_manual(values = site_colors_ID) +
  scale_x_discrete(labels = c(
    "Harry's Bommie" = "HB",
    "Fourth Point" = "FP",
    "Deep Lagoon" = "DL",
    "Shallow Lagoon" = "SL",
    "Reef Crest" = "RC",
    "Reef Flat" = "RF")) +
  theme_classic() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.title.x = element_blank(),
    panel.background = element_rect(fill = NA, color = "black"),
    legend.position = "none", legend.text = element_text(size = 16),
legend.title = element_text(size = 18)) + labs(y = "Distance to centroid")

phys_dist

# To assess within-site variability, we analyzed the dispersion of individuals in multivariate space. Significant differences in multivariate dispersion by site was observed (F=4.07, p=0.004), with SL exhibiting the greatest spread, indicating greater physiological heterogeneity within this site. Pairwise comparisons confirmed that SL was significantly more dispersed than FP, DL, HB, and RF (p<0.037). Differences in dispersion between other sites were not statistically significant (p>0.58).
```

##Inset PCA - environmental parameters
```{r}

env_vars <- c(
  "mean_o2",
  "mean_salinity",
  "mean_turbidity",
  "mean_depth",
  "mean.seasonal.amp",
  "mean.daily.amp",
  "DLI",
  "Fluor",
  "O2_amp")

env_fit_df <- phys_clean %>%
  dplyr::filter(Origin != "RC") %>%
  dplyr::select(Origin, Coral.ID, all_of(env_vars))


phys_scores_all <- as.data.frame(scores(phys.rda, display = "sites", scaling = 1))
phys_scores_all <- data.frame(phys_scores_all, phys_clean)

phys_scores_envfit <- phys_scores_all %>%
  dplyr::filter(Origin != "RC")

fit_env <- envfit(
  phys_scores_envfit[, c("PC1", "PC2")],
  env_fit_df[, env_vars],
  permutations = 999)

fit_env

env_scores <- as.data.frame(scores(fit_env, display = "vectors"))
env_scores$var  <- rownames(env_scores)
env_scores$pval <- fit_env$vectors$pvals
env_scores$r    <- fit_env$vectors$r

colnames(env_scores)[1:2] <- c("PC1", "PC2")

```

```{r}
env_scores$label <- dplyr::recode(env_scores$var,
  mean_o2 = "DO",
  mean_salinity = "Salinity",
  mean_turbidity = "Turbidity",
  mean_depth = "Depth",
  mean.seasonal.amp = "Mean~seasonal~amp.",
  mean.daily.amp = "Mean~daily~amp.",
  DLI = "PAR~daily~dose", 
  Fluor = "Fluor.", 
  O2_amp = "DO~amp.")

env_inset <- ggplot() +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.5, color = "gray50") +
  geom_vline(xintercept = 0, linetype = 2, alpha = 0.5, color = "gray50") +
  geom_segment(
    data = env_scores,
    aes(x = 0, y = 0, xend = PC1 * 1.4, yend = PC2 * 1.4),
    arrow = arrow(length = unit(0.2, "lines")),
    color = "black", alpha = 0.9) +
  geom_text_repel(data = env_scores,aes(x = PC1 * 1.4, y = PC2 * 1.4, label = label),size = 5,color = "black",box.padding = 0.25,point.padding = 0.15,force = 5,force_pull = 0.2,max.overlaps = Inf,min.segment.length = 0,segment.size = 0.2,seed = 123, parse = TRUE ) +
  scale_color_manual(values = site_colors_ID) +
  theme_classic() +
  theme(
    plot.margin = margin(0, 0, 0, 0),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = "black"),
    panel.border = element_rect(color = "black", fill = NA),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none")

env_inset

```

#FIGURE 3
```{r}
g <- g +coord_cartesian(xlim = c(-1.5, 2.5), ylim = c(-1.5, 2.5), clip = "off")

physfinal_pca <- g +
  annotation_custom(
    grob = ggplotGrob(env_inset),
    xmin = -1.71,
    xmax = -0.2,
    ymin = 1.2,
    ymax = 2.70)

physfinal_pca
```